---
title: "Advanced Workflows. Meta-Analysis, Longitudinal Data, Two-Part Models and Validation"
author: "Walter Esfrain Pereira"
output: rmarkdown::html_vignette
bibliography: references.bib
link-citations: true
vignette: >
  %\VignetteIndexEntry{Advanced Workflows. Meta-Analysis, Longitudinal Data, Two-Part Models and Validation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 6, fig.height = 3.6)
library(gamlssPosthoc)
has_core <- requireNamespace("gamlss", quietly = TRUE) &&
  requireNamespace("gamlss.dist", quietly = TRUE)
has_dist <- has_core && requireNamespace("distributions3", quietly = TRUE)
has_zadj <- has_dist && requireNamespace("gamlss.inf", quietly = TRUE)
has_gg   <- has_core && requireNamespace("ggplot2", quietly = TRUE)
has_nlme <- has_core && requireNamespace("nlme", quietly = TRUE)
if (has_core) suppressPackageStartupMessages(library(gamlss.dist))
```

# Purpose

This vignette covers the workflows where the scientific structure matters more
than a single comparison of means. Every section states the estimator, gives a
short executable demonstration, and ends with the failure mode that most often
invalidates the analysis in practice.

Sample sizes and replication counts here are deliberately small so that the
vignette builds quickly. Production values are indicated in each section. The
full-size scripts live in `inst/examples/`.

# 1. Meta-analysis of a GAMLSS estimand

## 1.1 Estimator

Fit the appropriate model in each study, extract the *same* estimand under the
*same* target population, and synthesize the resulting effects. With study
effects \(\hat\delta_s\) and standard errors \(SE_s\), the fixed-effect synthesis
uses inverse-variance weights,

\[
w_s = SE_s^{-2}, \qquad
\hat\delta_{\mathrm{FE}} = \frac{\sum_s w_s \hat\delta_s}{\sum_s w_s},
\qquad
\operatorname{Var}(\hat\delta_{\mathrm{FE}}) = \frac{1}{\sum_s w_s}.
\tag{1}
\]

Equation (1) assumes a single common effect. When studies differ in soil,
cultivar, or season that assumption is untenable, and the random-effects model
is the appropriate default,

\[
\hat\delta_s = \delta + u_s + e_s,
\qquad u_s \sim N(0, \tau^2),
\qquad e_s \sim N(0, SE_s^2).
\tag{2}
\]

A simple moment estimator of the between-study variance is the DerSimonian and
Laird form,

\[
\hat\tau^2 = \max\left\{0,\;
\frac{Q - (S-1)}{\sum_s w_s - \left(\sum_s w_s^2 \big/ \sum_s w_s\right)}
\right\},
\qquad
Q = \sum_s w_s\left(\hat\delta_s - \hat\delta_{\mathrm{FE}}\right)^2 ,
\tag{3}
\]

with the heterogeneity share

\[
I^2 = \max\left\{0,\; \frac{Q - (S-1)}{Q}\right\}.
\]

Weights are then recomputed as \(w_s^{*} = (SE_s^2 + \hat\tau^2)^{-1}\). For real
work use `metafor` [@viechtbauer2010], which provides REML estimation of
\(\tau^2\), prediction intervals, and diagnostics; the code below is a
transparent illustration, not a replacement.

## 1.2 Demonstration

```{r meta, eval=has_dist}
set.seed(5001)
sim_study <- function(s, effect) {
  d <- data.frame(trt = factor(rep(c("Control", "Bio"), each = 30)))
  d$y <- gamlss.dist::rGA(nrow(d),
    mu = exp(0.7 + effect * (d$trt == "Bio")), sigma = 0.30)
  fit <- gamlss::gamlss(y ~ trt, family = gamlss.dist::GA, data = d,
                        trace = FALSE)
  # The log-ratio is the scale to pool: symmetric and near-normal
  ph <- gamlss_posthoc(fit, "trt", estimand = "mean", contrast = "pairwise",
                       comparison = "log_ratio", engine = "distribution",
                       uncertainty = "bootstrap", B = 49L, data = d)
  c(delta = ph$contrasts$estimate[1], se = ph$contrasts$SE[1])
}
effects <- c(0.30, 0.22, 0.38)          # true study-specific effects
res <- t(vapply(seq_along(effects), function(s) sim_study(s, effects[s]),
                numeric(2)))
res <- as.data.frame(res)
res$study <- paste0("S", seq_len(nrow(res)))
res
```

```{r meta-synthesis, eval=has_dist}
w  <- 1 / res$se^2
fe <- sum(w * res$delta) / sum(w)
Q  <- sum(w * (res$delta - fe)^2)
S  <- nrow(res)
tau2 <- max(0, (Q - (S - 1)) / (sum(w) - sum(w^2) / sum(w)))
I2   <- max(0, (Q - (S - 1)) / Q)
wr <- 1 / (res$se^2 + tau2)
re <- sum(wr * res$delta) / sum(wr)
data.frame(fixed_effect = fe, random_effect = re,
           tau2 = tau2, I2 = I2,
           ratio_scale = exp(re))
```

## 1.3 Failure mode

The synthesis is only meaningful if every \(\hat\delta_s\) is the same estimand.
Pooling a difference of means from one study with a difference of linear
predictors from another produces a number with no interpretation. Record
`estimand_info` from each study alongside the effect, and refuse to pool when the
target populations differ.

# 2. Longitudinal data and the marginal-conditional gap

## 2.1 The gap

`gamlss` documents `random()` and `re()` for random-effect structures. `random()`
acts as a smoother for random intercepts; `re()` interfaces with `nlme::lme()`
for richer structures [@stasinopoulos2017].

With a random intercept \(b_j \sim N(0,\sigma_b^2)\) and a log link, the
subject-specific and population-averaged means differ by a fixed factor,

\[
\mathrm{E}\!\left[Y \mid b_j\right] = \exp(\eta + b_j),
\qquad
\mathrm{E}[Y] = \mathrm{E}_b\!\left[\exp(\eta + b)\right]
= \exp(\eta)\exp\!\left(\tfrac{1}{2}\sigma_b^2\right).
\tag{4}
\]

The attenuation factor \(\exp(\sigma_b^2/2)\) cancels in a *ratio* of two levels
but not in a *difference*. That is the practical reason to prefer the log-ratio
scale in mixed GAMLSS work, and the reason a marginal difference and a
conditional difference are two different estimands rather than two estimates of
one.

```{r gap, eval=TRUE}
sigma_b <- c(0.1, 0.3, 0.5, 0.8)
data.frame(sigma_b = sigma_b,
           attenuation = round(exp(sigma_b^2 / 2), 4))
```

## 2.2 Demonstration

```{r longitudinal, eval=has_nlme && has_dist}
set.seed(5002)
J <- 20; T_ <- 4
dl <- expand.grid(time = seq_len(T_), plot = factor(seq_len(J)))
dl$trt <- factor(ifelse(as.integer(dl$plot) %% 2 == 0, "Bio", "Control"),
                 levels = c("Control", "Bio"))
b <- stats::rnorm(J, 0, 0.30)
eta <- 0.6 + 0.25 * (dl$trt == "Bio") + 0.05 * dl$time + b[as.integer(dl$plot)]
dl$y <- gamlss.dist::rGA(nrow(dl), mu = exp(eta), sigma = 0.25)

fl <- gamlss::gamlss(y ~ trt + time + gamlss::random(plot),
                     family = gamlss.dist::GA, data = dl, trace = FALSE)
gamlss_posthoc(fl, "trt", estimand = "mean", contrast = "pairwise",
               comparison = "log_ratio", uncertainty = "none", data = dl)$contrasts
```

## 2.3 Failure mode

Reporting a subject-specific effect and describing it as a population effect. If
the question is "what happens to the field", the estimand must be marginal; if it
is "what happens to this plot", it must be conditional. Both are legitimate, and
they are numerically different under any non-identity link.

# 3. Hurdle, zero-adjusted and censored responses

## 3.1 Three models, one data pattern

A continuous positive response with exact zeros admits at least three
representations, and they are not interchangeable.

The **zero-adjusted** representation is a two-part mixture,

\[
f(y) = \begin{cases}
h, & y = 0,\\[2pt]
(1-h)\, f_{+}(y), & y > 0,
\end{cases}
\qquad
\mathrm{E}(Y) = (1-h)\,\mathrm{E}(Y_+).
\tag{5}
\]

The **hurdle** representation writes the same density with a truncated positive
part, \(f_{+}(y) = f(y) / \{1 - F(0)\}\). When the positive component is the same
family and the zero probability is modelled separately, the two are
reparameterizations of one another and give identical fitted densities. They
diverge as soon as the positive component is *not* truncated at zero in one of
them, or as soon as a shared parameter links the two parts.

The **censoring** representation assumes the zeros are not real zeros but values
below a detection limit \(L\),

\[
P(Y \le L) = F(L), \qquad \text{observed } y_i = \max(Y_i, L).
\]

The decision rule is scientific, not statistical:

| Origin of the zeros | Appropriate model |
|---|---|
| A structural process really produces exact zeros, such as complete crop failure | zero-adjusted or hurdle |
| The measurement device cannot read below a threshold | left-censored |
| Zeros are recording errors or missing values coded as zero | neither; treat as missing |

Choosing a two-part model for detection-limit zeros inflates the estimated zero
probability and biases the marginal mean downward. No goodness-of-fit statistic
will reveal that, because both models can fit the observed histogram.

## 3.2 Demonstration

```{r zadj, eval=has_zadj}
set.seed(5003)
dz <- data.frame(trt = factor(rep(c("Control", "Bio"), each = 45)))
h_true <- ifelse(dz$trt == "Control", 0.40, 0.18)
pos <- gamlss.dist::rGA(nrow(dz),
  mu = ifelse(dz$trt == "Control", 2.0, 3.0), sigma = 0.30)
dz$y <- ifelse(stats::rbinom(nrow(dz), 1, h_true) == 1, 0, pos)

zfit <- gamlss.inf::gamlssZadj(y = y, mu.formula = ~ trt, xi0.formula = ~ trt,
                               family = gamlss.dist::GA, data = dz,
                               trace = FALSE)
nd <- data.frame(trt = factor(levels(dz$trt), levels = levels(dz$trt)))
gamlss_distribution_summary(zfit, nd, data = dz)
```

The three scientific components, the zero probability, the conditional positive
mean and the marginal mean, are separable and should be reported together.

```{r zadj-plot, eval=has_gg && has_zadj}
plot_gamlss_zero_adjusted(zfit, "trt", data = dz)
```

## 3.3 Failure mode

Comparing only the positive part and calling it the treatment effect. If the
treatment works by reducing the probability of total failure, the positive-part
comparison can be flat while the marginal mean moves substantially.

# 4. Cross-validation with proper scoring rules

## 4.1 Why a proper rule

A scoring rule \(S(F, y)\) is *proper* when the expected score is optimized by
the true distribution, and *strictly proper* when that optimum is unique
[@gneiting2007]. Optimizing an improper rule rewards a forecaster for reporting
something other than their honest belief, which is exactly what a model-selection
criterion must not do.

For a distributional regression, the mean squared error is not the right
criterion, because it is blind to everything except the first moment. A GAMLSS
that halves the predictive dispersion at constant mean scores identically under
MSE and much better under any strictly proper rule.

Two strictly proper rules are used here. The **logarithmic score**,

\[
LS_i = -\log f_i(y_i),
\tag{6}
\]

is local and severe: a single observation in a region of near-zero predictive
density dominates the average. For a zero-adjusted model the density in equation
(6) is the point mass \(h_i\) when \(y_i = 0\) and \((1 - h_i) f_{+}(y_i)\)
otherwise.

The **continuous ranked probability score**,

\[
CRPS_i = \int_{-\infty}^{\infty}\left\{F_i(z) - \mathbb{1}(y_i \le z)\right\}^2 dz,
\tag{7}
\]

is also strictly proper, is expressed in the units of the response, and is far
more robust than equation (6) to a single badly placed observation. For skewed
agronomic responses CRPS is usually the better default.

## 4.2 Demonstration

```{r cv, eval=has_dist}
set.seed(5004)
dcv <- data.frame(dose = rep(seq(0, 100, length.out = 25), 4))
dcv$y <- gamlss.dist::rGA(nrow(dcv),
  mu = exp(0.8 + 0.008 * dcv$dose), sigma = 0.35)

crps_sample <- function(F_draws, y) {
  # Empirical CRPS from predictive draws, using the standard identity
  # CRPS = E|X - y| - 0.5 E|X - X'|
  a <- mean(abs(F_draws - y))
  b <- mean(abs(outer(F_draws, F_draws, "-"))) / 2
  a - b
}

folds <- rep(1:4, length.out = nrow(dcv))
score <- function(family) {
  out <- vapply(1:4, function(k) {
    tr <- dcv[folds != k, ]; te <- dcv[folds == k, ]
    f <- try(gamlss::gamlss(y ~ dose, family = family, data = tr, trace = FALSE),
             silent = TRUE)
    if (inherits(f, "try-error")) return(c(NA_real_, NA_real_))
    mu <- stats::predict(f, what = "mu", newdata = te, type = "response", data = tr)
    sg <- stats::predict(f, what = "sigma", newdata = te, type = "response", data = tr)
    if (length(sg) == 1L) sg <- rep(sg, nrow(te))
    dens <- if (identical(family, "GA")) {
      gamlss.dist::dGA(te$y, mu = mu, sigma = sg)
    } else {
      gamlss.dist::dNO(te$y, mu = mu, sigma = sg)
    }
    ls <- mean(-log(pmax(dens, 1e-300)))
    cr <- mean(vapply(seq_len(nrow(te)), function(i) {
      dr <- if (identical(family, "GA")) {
        gamlss.dist::rGA(200, mu = mu[i], sigma = sg[i])
      } else {
        gamlss.dist::rNO(200, mu = mu[i], sigma = sg[i])
      }
      crps_sample(dr, te$y[i])
    }, numeric(1)))
    c(ls, cr)
  }, numeric(2))
  c(log_score = mean(out[1, ], na.rm = TRUE),
    crps = mean(out[2, ], na.rm = TRUE))
}
rbind(GA = score("GA"), NO = score("NO"))
```

Lower is better for both rules. The Gamma family, which generated the data,
should win on both; when the two rules disagree, the log score is usually being
driven by one extreme observation and CRPS is the more stable guide.

For Bayesian fits the leave-one-out machinery of `loo` provides the same
logarithmic score with importance-sampling diagnostics [@vehtari2017]; the
frequentist k-fold loop above is the analogue available here.

## 4.3 Failure mode

Selecting the family by AIC alone. AIC is an in-sample criterion tied to the
logarithmic score and to a fixed penalty; it does not test predictive
calibration of the dispersion or shape parameters, which is precisely what a
distributional model claims to get right.

# 5. Computational cost

Uncertainty layers dominate the run time, not the engines. The table gives the
number of model fits required for one call.

| Layer | Model refits | Typical relative cost |
|---|---|---|
| `none` | 1 | 1 |
| `delta` | 1 | 1 to 2 |
| `simulation` | 1, plus `B` coefficient draws | 2 to 5 |
| `bootstrap` | `1 + B` | `B` |

With `B = 499`, a refit bootstrap is roughly five hundred times the cost of a
point estimate. Two practical rules follow. Prototype with `uncertainty = "none"`
and switch on the bootstrap once the estimand is settled. And when a bootstrap is
unavoidable, use `parallel = TRUE` with `cores` set below the physical core
count, since each GAMLSS refit is already partly vectorized.

`tools/benchmark.R` in the source tree compares engine and bootstrap costs.
Benchmarks are a development tool and are not part of the API executed at load
time.

# References
