---
title: "A Complete Agronomic Case Study from Design to Results Paragraph"
author: "Walter Esfrain Pereira"
output: rmarkdown::html_vignette
bibliography: references.bib
link-citations: true
vignette: >
  %\VignetteIndexEntry{A Complete Agronomic Case Study from Design to Results Paragraph}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 6.2, fig.height = 3.8)
library(gamlssPosthoc)
has_core <- requireNamespace("gamlss", quietly = TRUE) &&
  requireNamespace("gamlss.dist", quietly = TRUE)
has_dist <- has_core && requireNamespace("distributions3", quietly = TRUE)
has_gg   <- has_core && requireNamespace("ggplot2", quietly = TRUE)
has_emm  <- has_core && requireNamespace("emmeans", quietly = TRUE)
has_mcv  <- has_core && requireNamespace("multcompView", quietly = TRUE)
if (has_core) suppressPackageStartupMessages(library(gamlss.dist))
```

# 1. The trial and the question

A randomized complete block experiment evaluates four nitrogen sources on grain
yield, in five blocks, with three plots per source per block. Yield is a positive
continuous response, right-skewed, and field experience says the sources differ
not only in average yield but in **consistency**: one of them is known to perform
well on good plots and poorly on marginal ones.

Two questions follow, and they need two different estimands.

1. Does mean yield differ among sources? The estimand is the standardized
   response mean, compared pairwise, on the ratio scale because agronomic
   recommendations are relative.
2. Does yield *stability* differ? The estimand is the standardized variance, or
   equivalently the dispersion parameter, compared on the same scale.

A model that estimates only the mean cannot answer the second question. That is
the reason to fit a GAMLSS rather than a linear model [@rigby2005].

The data below are simulated so the vignette is self-contained and reproducible.
The generating process is written out explicitly so the reader can check what the
analysis recovers.

```{r data, eval=has_core}
set.seed(7001)
sources <- c("Urea", "AmmSulf", "CoatedUrea", "Organic")
d <- expand.grid(
  rep    = 1:3,
  nsource = factor(sources, levels = sources),
  block  = factor(paste0("B", 1:5))
)
d$rep <- NULL

# Mean structure: block effect plus source effect, multiplicative
mu_source <- c(Urea = 0.00, AmmSulf = 0.06, CoatedUrea = 0.18, Organic = -0.05)
mu_block  <- c(B1 = -0.10, B2 = 0.00, B3 = 0.05, B4 = 0.12, B5 = -0.07)
eta_mu <- log(4.2) + mu_source[as.character(d$nsource)] +
                     mu_block[as.character(d$block)]

# Dispersion structure: Organic is markedly less stable
sg_source <- c(Urea = 0.00, AmmSulf = -0.05, CoatedUrea = -0.12, Organic = 0.45)
eta_sg <- log(0.16) + sg_source[as.character(d$nsource)]

d$yield <- gamlss.dist::rGA(nrow(d), mu = exp(eta_mu), sigma = exp(eta_sg))
str(d)
```

One naming detail is worth flagging, because it costs hours when it bites. The
factor is called `nsource` and not `source`. Internally `predict.gamlss()` binds
the original data and the prediction data into one frame and marks the origin of
each row with a column literally named `source`. A covariate of that name is
silently overwritten, and prediction then fails with an opaque
`lm.wfit ... 0 non-NA cases` error. The same applies to any other name the
internal helper uses. When a model refuses to predict for no visible reason,
renaming the covariate is the first thing to try.

# 2. Choosing the family by diagnosis, not by habit

The default reflex is Gaussian. For a positive, right-skewed response with
multiplicative block effects that is rarely the right choice. Compare candidates
on out-of-sample behaviour and on residual diagnostics, not on convenience.

```{r family, eval=has_core}
f_no <- gamlss::gamlss(yield ~ nsource + block, family = gamlss.dist::NO,
                       data = d, trace = FALSE)
f_ga <- gamlss::gamlss(yield ~ nsource + block, sigma.formula = ~ nsource,
                       family = gamlss.dist::GA, data = d, trace = FALSE)
f_ln <- gamlss::gamlss(yield ~ nsource + block, sigma.formula = ~ nsource,
                       family = gamlss.dist::LOGNO, data = d, trace = FALSE)
gamlss_poly_compare(Normal = f_no, Gamma = f_ga, Lognormal = f_ln)
```

Randomized quantile residuals [@dunn1996] are the right diagnostic here, because
they are exactly standard normal under a correctly specified model regardless of
the family, which ordinary Pearson residuals are not.

```{r diagnostics, eval=has_gg}
plot_gamlss_diagnostics(f_ga, "qq", data = d)
plot_gamlss_diagnostics(f_ga, "pit", bins = 10L, data = d)
```

A well-calibrated model gives a straight quantile-quantile plot and a flat
histogram of the probability integral transform. A U-shaped PIT histogram means
the predictive distribution is too narrow, and a hump-shaped one means it is too
wide. Neither is visible in the mean structure alone.

```{r fit, eval=has_core}
fit <- f_ga
fit
```

# 3. Declare the plan before computing any contrast

`gamlss_posthoc_plan()` is not decoration. It reports which engine is eligible,
which parameters were discovered, and which uncertainty layer is defensible for
the requested target. Running it first prevents the most common failure, which is
requesting a delta interval for a quantity that does not have one.

```{r plan, eval=has_core}
gamlss_posthoc_plan(fit, estimand = "mean", contrast = "pairwise",
                    comparison = "ratio")
```

# 4. Question one, mean yield

The estimand is fixed by four choices: the response mean, standardized over the
observed covariate distribution, compared pairwise, on the ratio scale. The
uncertainty layer is a refit bootstrap, because a marginal mean derived from the
full predictive distribution has no reliable joint covariance in current
`gamlss`.

```{r mean-contrasts, eval=has_dist}
ph_mean <- gamlss_posthoc(
  fit, "nsource", estimand = "mean",
  contrast = "pairwise", comparison = "ratio",
  population = "observed",
  uncertainty = "bootstrap", B = 99L, seed = 7002,
  data = d
)
ph_mean$estimand_info
est <- ph_mean$estimates[, intersect(c("nsource", "estimate", "SE",
                                       "lower.CL", "upper.CL"),
                                     names(ph_mean$estimates))]
num <- vapply(est, is.numeric, logical(1))
est[num] <- lapply(est[num], round, 4)
est
```

```{r mean-table, eval=has_dist}
cols <- intersect(c("comparison", "estimate", "lower.CL", "upper.CL", "p.value"),
                  names(ph_mean$contrasts))
round_df <- ph_mean$contrasts[, cols]
num <- vapply(round_df, is.numeric, logical(1))
round_df[num] <- lapply(round_df[num], round, 4)
round_df
```

```{r mean-plot, eval=has_gg && has_dist}
plot_gamlss_contrasts(ph_mean, null = 1) +
  ggplot2::labs(x = "Yield ratio versus the comparator",
                title = "Pairwise ratios of standardized mean yield")
```

# 5. Question two, stability

The same object-oriented workflow answers the dispersion question. Nothing
changes except the estimand.

```{r variance, eval=has_dist}
ph_var <- gamlss_posthoc(
  fit, "nsource", estimand = "variance",
  contrast = "reference", comparison = "ratio",
  uncertainty = "bootstrap", B = 99L, seed = 7003,
  data = d
)
cols <- intersect(c("comparison", "estimate", "lower.CL", "upper.CL"),
                  names(ph_var$contrasts))
z <- ph_var$contrasts[, cols]
num <- vapply(z, is.numeric, logical(1)); z[num] <- lapply(z[num], round, 4)
z
```

```{r params-plot, eval=has_gg && has_dist}
plot_gamlss_parameters(fit, "nsource", parameters = c("mu", "sigma"), data = d)
```

The two panels carry the whole scientific message: the coated source raises the
mean, and the organic source raises the dispersion. A mean-only analysis would
have reported the first and hidden the second.

# 6. Compact letter displays, and their limits

Letters are requested constantly by agronomic journals, so the package provides
them through the algorithm of Piepho [@piepho2004]. They require p-values, so the
object must carry an uncertainty layer.

```{r cld, eval=has_mcv && has_emm}
ph_letters <- gamlss_posthoc(
  fit, "nsource", estimand = "parameter", what = "mu",
  population = "reference", contrast = "pairwise",
  engine = "emmeans", uncertainty = "delta", data = d
)
gamlss_cld(ph_letters)
```

The `alpha` argument controls only the display threshold, not the inference. A
regression context where evidence up to `p = 0.10` is treated as meaningful is
declared by setting it explicitly:

```{r cld-10, eval=has_mcv && has_emm}
gamlss_cld(ph_letters, alpha = 0.10)
```

## 6.1 Why letters should be used with care

Three criticisms are well founded and should be acknowledged in any manuscript
that prints letters.

1. **Sharing a letter is not evidence of equality.** Two means share a letter
   whenever the test failed to reject, which happens both when they are genuinely
   close and when the experiment lacked power. The display cannot distinguish
   those two situations, and readers routinely misread it as proof of no
   difference.
2. **The display is threshold-dependent and unstable.** Moving `alpha` from 0.05
   to 0.10 can change the entire pattern, as the two blocks above show, even
   though no estimate moved at all.
3. **Effect size and uncertainty disappear.** A letter carries no information
   about how large a difference is or how precisely it was measured, which is
   exactly the information a reader needs to judge agronomic relevance.

The recommendation implemented throughout this package is therefore to report the
estimate with its interval as the primary result, and to treat letters as an
optional reading aid printed alongside, never as a replacement. The contrast plot
in section 4 conveys size, direction and precision at once; the letters do not.

# 7. Figures for submission

```{r figures, eval=has_gg && has_dist}
p1 <- plot_gamlss_estimand(fit, "nsource", estimand = "mean", data = d) +
  ggplot2::labs(y = "Standardized mean yield (t/ha)", x = "Nitrogen source") +
  theme_gamlss_journal()
p1
```

```{r figures2, eval=has_gg && has_dist}
plot_gamlss_distribution(fit, "nsource", type = "density", data = d,
                         n_response = 60L) +
  ggplot2::labs(x = "Yield (t/ha)", y = "Predictive density",
                title = "Standardized predictive distributions") +
  theme_gamlss_journal()
```

The predictive densities make the stability result visible without any test: the
organic source has a visibly wider distribution at a similar location.

# 8. A written Results paragraph

The point of recording the estimand is that the paragraph can be written from the
object, not from memory. Everything in italics below is read directly from
`estimand_info` and from the contrast table.

> Grain yield was modelled with a Gamma GAMLSS with nitrogen source and block in
> the location submodel and nitrogen source in the dispersion submodel
> (log links for both). Model adequacy was assessed with randomized quantile
> residuals. The reported estimand is the *ratio of response means, standardized
> over the observed covariate distribution*, with *pairwise* comparisons among
> sources and *99 case-refit bootstrap* replicates for interval estimation.
> Coated urea increased mean yield relative to urea (ratio and 95 percent
> interval reported in Table 1), while the organic source did not differ from
> urea in the mean. Yield dispersion differed: the standardized variance under
> the organic source was substantially higher than under urea (Table 2),
> indicating lower stability at comparable mean performance. Evidence up to
> `p = 0.10` was treated as meaningful, and compact letters are shown in Figure 1
> as a reading aid only; the primary inference is based on the reported effect
> sizes and intervals.

Three properties of that paragraph are worth naming. It states the estimand
before the result. It separates the mean conclusion from the dispersion
conclusion. And it says what the letters are for, which prevents the reader from
over-reading them.

# 9. Export

```{r export, eval=has_core}
tex <- export_to_latex(ph_mean, component = "contrasts",
                       caption = "Pairwise ratios of standardized mean yield")
cat(substr(tex, 1, 200))
```

```{r report, eval=has_dist}
md <- generate_report(ph_mean, output = "md",
                      title = "Nitrogen source trial", include_plots = FALSE)
cat(substr(md, 1, 200))
```

# 10. Reproducibility checklist

Before submission, confirm every line below.

| Item | Where it lives |
|---|---|
| Family chosen by diagnosis, with the comparison recorded | `gamlss_poly_compare()` output |
| Residual diagnostics inspected | `plot_gamlss_diagnostics()` |
| Engine and uncertainty route decided in advance | `gamlss_posthoc_plan()` |
| Estimand, population, geometry and scale stated | `ans$estimand_info` |
| Bootstrap seed fixed | `seed` argument |
| Uncertainty layer named in the text | `ans$notes` |
| Letters, if printed, described as a reading aid | manuscript text |
| Session information archived | `sessionInfo()` |

```{r session, eval=TRUE}
utils::sessionInfo()$R.version$version.string
```

# References
