---
title: "Standardization and the Choice of Target Population"
author: "Walter Esfrain Pereira"
output: rmarkdown::html_vignette
bibliography: references.bib
link-citations: true
vignette: >
  %\VignetteIndexEntry{Standardization and the Choice of Target Population}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 6, fig.height = 3.6)
library(gamlssPosthoc)
has_core <- requireNamespace("gamlss", quietly = TRUE) &&
  requireNamespace("gamlss.dist", quietly = TRUE)
has_dist <- has_core && requireNamespace("distributions3", quietly = TRUE)
has_gg   <- has_core && requireNamespace("ggplot2", quietly = TRUE)
has_emm  <- has_core && requireNamespace("emmeans", quietly = TRUE)
if (has_core) suppressPackageStartupMessages(library(gamlss.dist))
```

# 1. The question this vignette answers

Two analysts fit the same GAMLSS to the same data, request "the difference
between treatments", and obtain different numbers. Neither made a mistake. They
answered different questions, because they averaged over different populations.

This vignette makes that choice explicit, gives the three weighting rules the
package implements, and quantifies how much a published conclusion can move
between them.

# 2. Standardization as a plug-in of the g-formula

Let \(A\) denote the focal factor, \(\mathbf{X}\) the remaining covariates, and
\(Y^{a}\) the response that would be observed under level \(a\). Under the usual
identification conditions of conditional exchangeability, positivity, and
consistency, the marginal distribution of the potential outcome is recovered by
averaging the conditional model over the covariate distribution [@robins1986],

\[
P\!\left(Y^{a} \le y\right)
= \int P\!\left(Y \le y \mid A = a, \mathbf{X} = \mathbf{x}\right)\,
  dP(\mathbf{x}).
\tag{1}
\]

A GAMLSS supplies the inner conditional distribution in closed form, since it
models every distributional parameter. Replacing \(dP(\mathbf{x})\) by a discrete
weighting \(\{w_i\}\) over the sample yields the plug-in estimator that this
package computes,

\[
\widehat{F}_a(y) = \sum_{i=1}^{n} w_i\,
\mathcal{D}\!\left(y \;\middle|\; \hat\theta_1(\mathbf{X}_i, a), \ldots,
\hat\theta_p(\mathbf{X}_i, a)\right),
\qquad \sum_i w_i = 1 .
\tag{2}
\]

Any reported estimand is then \(T(\widehat{F}_a)\) for a functional \(T\).

Two remarks matter in practice. Equation (2) is a *model-based*
standardization, so its validity rests on the GAMLSS being adequate for the
conditional distribution and not only for its mean. And the word "causal" is
earned only when the identification conditions hold by design; in an
observational study equation (2) is still a well-defined descriptive summary,
but it is not a causal contrast.

# 3. The three weighting rules

The package exposes three target populations. They differ only in \(w_i\) and in
which rows enter the sum.

| `population` | Rows used | Weight \(w_i\) | Question answered |
|---|---|---|---|
| `"observed"` | all \(n\) rows, with the focal factor set to \(a\) | \(w_i = 1/n\), or user weights | What would the average response be if this same population received \(a\)? |
| `"balanced"` | all \(n\) rows | \(w_i \propto 1 / n_{c(i)}\), with \(n_{c}\) the size of the covariate cell of row \(i\) | What would it be in a population where the covariate cells are equally frequent? |
| `"reference"` | one row per covariate combination | \(w_i = 1 / K\) over the \(K\) grid cells | What is it at a constructed reference profile? |

Formally, writing \(c(i)\) for the covariate cell of row \(i\),

\[
w_i^{\text{obs}} = \frac{1}{n}, \qquad
w_i^{\text{bal}} = \frac{1/n_{c(i)}}{\sum_{j} 1/n_{c(j)}}, \qquad
w_i^{\text{ref}} = \frac{1}{K}\,\mathbb{1}\{i \in \mathcal{G}\}.
\]

The observed population is the default because it is the only one that requires
no assumption beyond the fitted model. The balanced population answers a
counterfactual demographic question and should be declared as such. The
reference population is a *profile*, not a population, and is the only one for
which `emmeans` is automatically eligible in this package, precisely because a
reference grid is what `emmeans` constructs [@lenth_emmeans].

# 4. A design where the three disagree

The three rules coincide only when the covariate distribution is the same in
every treatment group and the link is the identity. The example below violates
both conditions on purpose: the covariate `soil` is unbalanced across
treatments, and the family has a log link.

```{r data, eval=has_core}
set.seed(4101)
n_cell <- c("Control.Sandy" = 40, "Control.Clay" = 10,
            "Bio.Sandy"     = 10, "Bio.Clay"     = 40)
cells <- do.call(rbind, lapply(names(n_cell), function(k) {
  parts <- strsplit(k, ".", fixed = TRUE)[[1]]
  data.frame(trt = parts[1], soil = parts[2], n = n_cell[[k]])
}))
d <- cells[rep(seq_len(nrow(cells)), cells$n), c("trt", "soil")]
d$trt  <- factor(d$trt,  levels = c("Control", "Bio"))
d$soil <- factor(d$soil, levels = c("Sandy", "Clay"))
rownames(d) <- NULL

# Clay soil is more productive, and the treatment effect is multiplicative
eta <- 0.55 + 0.28 * (d$trt == "Bio") + 0.45 * (d$soil == "Clay")
d$y  <- gamlss.dist::rGA(nrow(d), mu = exp(eta), sigma = 0.32)
table(d$trt, d$soil)
```

The design is strongly unbalanced: the treated group sits mostly on the more
productive soil. A naive group mean would attribute the soil advantage to the
treatment.

```{r fit, eval=has_core}
fit <- gamlss::gamlss(y ~ trt + soil, sigma.formula = ~ soil,
                      family = gamlss.dist::GA, data = d, trace = FALSE)
# The raw group means are confounded by soil
tapply(d$y, d$trt, mean)
```

# 5. The same contrast under the three populations

```{r three-populations, eval=has_dist}
pops <- c("observed", "balanced")
rows <- lapply(pops, function(p) {
  z <- gamlss_posthoc(fit, "trt", estimand = "mean", contrast = "pairwise",
                      comparison = "difference", population = p,
                      uncertainty = "none", data = d)
  data.frame(population = p, comparison = z$contrasts$comparison,
             estimate = round(z$contrasts$estimate, 4))
})
if (has_emm) {
  zr <- gamlss_posthoc(fit, "trt", estimand = "parameter", what = "mu",
                       contrast = "pairwise", population = "reference",
                       engine = "emmeans", uncertainty = "none", data = d)
  rows[[length(rows) + 1L]] <- data.frame(
    population = "reference (mu, link scale)",
    comparison = as.character(zr$contrasts[[1]]),
    estimate = round(zr$contrasts$estimate, 4))
}
do.call(rbind, rows)
```

The observed and balanced estimates are both defensible, and they are not the
same number. The reference row is deliberately labelled with its scale, because
an `emmeans` contrast of `mu` under a log link is a difference of linear
predictors, that is, a log-ratio, and not a difference of means.

# 6. Why the ratio of averages is not the average of ratios

For a multiplicative model the ratio scale is often the natural one. It is
tempting to compute a ratio for each observation and average them. That is a
different estimand:

\[
\frac{\sum_i w_i \bar{T}_{a}(\mathbf{X}_i)}{\sum_i w_i \bar{T}_{b}(\mathbf{X}_i)}
\;\ne\;
\sum_i w_i \frac{\bar{T}_{a}(\mathbf{X}_i)}{\bar{T}_{b}(\mathbf{X}_i)}
\]

unless the row-wise ratios are constant. The package computes the left-hand
side, a ratio of standardized estimands, and records the choice in
`estimand_info`.

```{r ratio-of-averages, eval=has_dist}
nd_bio <- transform(d, trt = factor("Bio", levels = levels(d$trt)))
nd_ctl <- transform(d, trt = factor("Control", levels = levels(d$trt)))
p_bio <- stats::predict(fit, what = "mu", newdata = nd_bio,
                        type = "response", data = d)
p_ctl <- stats::predict(fit, what = "mu", newdata = nd_ctl,
                        type = "response", data = d)

pkg <- gamlss_posthoc(fit, "trt", estimand = "mean", contrast = "pairwise",
                      comparison = "ratio", uncertainty = "none",
                      data = d)$contrasts$estimate

data.frame(
  quantity = c("ratio of standardized means (package)",
               "average of row-wise ratios"),
  value = round(c(pkg, mean(p_bio / p_ctl)), 5)
)
```

In this particular model the two happen to agree closely because the treatment
enters additively on the log scale with no interaction. Add a `trt:soil`
interaction and they separate; the point is that the equality is a special case,
not a rule, and only one of the two is the standardized estimand.

# 7. Weights supplied by the analyst

Numeric weights are accepted directly and are normalized internally. This covers
survey weights, plot area, and any external standard population.

```{r weights, eval=has_dist}
# Give clay plots twice the weight of sandy plots, as an external standard
w <- ifelse(d$soil == "Clay", 2, 1)
gamlss_posthoc(fit, "trt", estimand = "mean", contrast = "pairwise",
               comparison = "difference", weights = w,
               uncertainty = "none", data = d)$contrasts
```

One warning is built into the package and deserves repeating here. Under a
weighted population, a non-linear comparison such as a ratio or a percent change
cannot be delegated to `marginaleffects::avg_comparisons()` without changing its
meaning, so that route is disabled and the distribution engine is used instead.

# 8. Standardization applies to the whole distribution

Nothing in equation (2) is specific to the mean. The same weights produce a
standardized variance, a standardized quantile, and a standardized exact zero
mass. This is where a distributional model earns its keep: a treatment can leave
the mean untouched and still halve the dispersion.

```{r whole-distribution, eval=has_dist}
targets <- c("mean", "variance")
do.call(rbind, lapply(targets, function(tg) {
  z <- gamlss_posthoc(fit, "trt", estimand = tg, contrast = "pairwise",
                      comparison = "ratio", uncertainty = "none", data = d)
  data.frame(estimand = tg, comparison = z$contrasts$comparison,
             ratio = round(z$contrasts$estimate, 4))
}))
```

```{r dist-picture, eval=has_gg && has_dist}
plot_gamlss_distribution(fit, "trt", type = "density", data = d,
                         n_response = 60L) +
  ggplot2::labs(subtitle = "Standardized over the observed covariate distribution")
```

# 9. Reporting checklist

A methods sentence that pins the estimand down completely needs five items, and
all five are stored in the returned object:

```{r report, eval=has_dist}
ans <- gamlss_posthoc(fit, "trt", estimand = "mean", contrast = "pairwise",
                      comparison = "ratio", population = "observed",
                      uncertainty = "none", data = d)
ans$estimand_info
ans$notes
```

Written out, that becomes: *the estimand is the ratio of response means,
standardized over the observed covariate distribution, comparing Bio with
Control, reported as a point estimate without an uncertainty layer.* A reader
can reproduce that sentence without seeing the code.

# 10. Limitations

- Equation (2) is model-based. A misspecified conditional distribution biases
  the standardized estimand even when the sample is large.
- Positivity is a real constraint. If a covariate cell contains no observations
  under one treatment level, the balanced and reference populations extrapolate,
  and the package cannot detect that for you.
- Under `population = "reference"` with a non-identity link, an `emmeans`
  contrast lives on the link scale unless `regrid()` is applied. The package
  applies `regrid(transform = "response")` when it can, and records the result in
  `notes`.

# References
