---
title: "Foundations, functions and examples of gamlssPosthoc"
author: "Walter Esfrain Pereira"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Foundations, functions and examples of gamlssPosthoc}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>",
                      fig.width = 7, fig.height = 4.5)
library(gamlssPosthoc)
has_core <- requireNamespace("gamlss", quietly = TRUE) &&
  requireNamespace("gamlss.dist", quietly = TRUE)
has_dist <- has_core && requireNamespace("distributions3", quietly = TRUE)
has_zadj <- has_dist && requireNamespace("gamlss.inf", quietly = TRUE)
has_emm  <- has_core && requireNamespace("emmeans", quietly = TRUE)
has_cld  <- has_emm && requireNamespace("multcompView", quietly = TRUE)
if (has_core) {
  suppressMessages(library(gamlss))
  suppressMessages(library(gamlss.dist))
}
if (has_zadj) suppressMessages(library(gamlss.inf))
```

**Reference version:** `gamlssPosthoc 0.2.0`

---

## Contents

1. Introduction
2. Why `gamlssPosthoc` was written
3. Statistical foundations
4. Conceptual architecture of the package
5. Installation and dependencies
6. `gamlss_posthoc()`
7. `gamlss_posthoc_plan()`
8. `gamlss_distribution_summary()`
9. `gamlss_trend()`
10. `gamlss_poly_compare()`
11. `gamlss_cld()`
12. `gamlss_engine_info()`
13. Recommended analysis workflows
14. Zero-adjusted models
15. Comparisons, multiplicity and interpretation
16. Quantitative regression, derivatives and optima
17. Uncertainty: delta, simulation and bootstrap
18. Good practice and limitations
19. Summary of the functions
20. References

---

# 1. Introduction

**Generalized Additive Models for Location, Scale and Shape (GAMLSS)** widen
the scope of generalized linear models considerably, because they allow more
than one parameter of the response distribution to be modelled as a function
of covariates. Instead of restricting inference to a conditional mean, a
GAMLSS model can describe location, dispersion, skewness, kurtosis or other
distribution-specific parameters at the same time (Rigby & Stasinopoulos,
2005; Stasinopoulos & Rigby, 2007; Stasinopoulos, Rigby & De Bastiani, 2018).

In general terms,

\[
Y_i \sim D(\theta_{1i},\theta_{2i},\ldots,\theta_{pi}),
\]

where each parameter may have its own predictor,

\[
g_k(\theta_{ki})
=
\eta_{ki}
=
\mathbf{x}_{ki}^{\mathsf T}\boldsymbol{\beta}_k
+
\sum_j f_{kj}(z_{kji}),
\qquad k=1,\ldots,p.
\]

In classical `gamlss`, the parameters are conventionally named `mu`, `sigma`,
`nu` and `tau`, although their interpretation depends on the chosen family.
Therefore, **`mu` must not be read automatically as the mean of the
response**. In some families `mu` is the mean; in others it is a median, a
location parameter, or something else entirely (Rigby et al., 2019).

That flexibility creates a post-modelling problem: classical
"mean comparison" procedures do not always correspond to the scientific
quantity the researcher intends to compare.

The `gamlssPosthoc` package was written to organise this stage. Its
philosophy is:

\[
\boxed{
\text{model}
\rightarrow
\text{predicted distribution}
\rightarrow
\text{estimand}
\rightarrow
\text{target population}
\rightarrow
\text{contrast}
\rightarrow
\text{uncertainty}
}
\]

Instead of asking only "what are the adjusted means?", the package forces a
sharper question:

> **Which quantity of the distribution is to be compared, for which
> population, and on which scientific scale?**

---

# 2. Why `gamlssPosthoc` was written

## 2.1. Limitations of a conventional post-hoc comparison

In simple linear models, comparing adjusted means is usually straightforward.
In GAMLSS, however, there are at least four difficulties.

### 2.1.1. The mean may depend on several parameters

Consider a distribution in which

\[
E(Y_i)=m(\mu_i,\sigma_i,\nu_i,\tau_i).
\]

Comparing `mu` alone need not be equivalent to comparing

\[
E(Y_i).
\]

The difference becomes even more important in zero-adjusted distributions,
where the marginal mean depends simultaneously on the continuous part and on
the mass at zero.

### 2.1.2. Support for `gamlss` in `emmeans` is partial

`emmeans` is a well-established tool for estimated marginal means and
contrasts (Lenth, 2016). For `gamlss` objects, however, its current support
is centred on one parameter at a time (`mu`, `sigma`, `nu` or `tau`), and it
does not cover the selected submodel when that submodel contains smoothers
such as `pb()`.

A model such as

```{r illustrative-01, eval=FALSE}
gamlss(
  y ~ treatment + pb(dose),
  sigma.formula = ~ treatment,
  family = GA,
  data = dat
)
```

should therefore not be forced through an EMM interface on the `mu` component
merely because that strategy works for `lm()` or `glm()`.

### 2.1.3. A "marginal mean" depends on the population it is marginalised over

A standardized mean can be defined over:

- the observed empirical distribution of the covariates;
- an artificial balanced population;
- a reference profile.

These estimands are not necessarily equal.

### 2.1.4. The scientific scale may differ from the link scale

With a log link, for instance, a natural contrast on the linear scale induces
ratios on the response scale. The researcher may nevertheless want,
explicitly,

\[
\Delta = m_A-m_B,
\]

\[
R = \frac{m_A}{m_B},
\]

\[
\log R = \log(m_A)-\log(m_B),
\]

or

\[
PC =
100\left(\frac{m_A}{m_B}-1\right).
\]

`gamlssPosthoc` keeps these choices separate.

---

# 3. Statistical foundations

## 3.1. GAMLSS as distributional regression

The GAMLSS principle is to model the complete conditional distribution of the
response. This perspective is often called **distributional regression**,
because the target is no longer only \(E(Y\mid X)\) but the whole set of
parameters of \(F(Y\mid X)\) (Stasinopoulos, Rigby & De Bastiani, 2018;
Umlauf, Klein & Zeileis, 2018).

For a generic parameter \(\theta_k\),

\[
g_k(\theta_{ki})
=
\eta_{ki}.
\]

The predictor may contain:

- linear effects;
- factors;
- interactions;
- polynomials;
- P-splines;
- random effects;
- other additive terms.

This structure makes it inadequate to assume that all post-model inference
can be reduced to a table of means.

---

## 3.2. Estimand

An **estimand** is the statistical quantity the analysis intends to estimate.

In `gamlssPosthoc`, the main estimands are:

```{r illustrative-02, eval=FALSE}
estimand = "parameter"
estimand = "mean"
estimand = "variance"
estimand = "quantile"
estimand = "prob_zero"
estimand = "custom"
```

### Distributional parameter

\[
T(F_i)=\theta_{ki}.
\]

For example, `what = "sigma"` requests inference on the dispersion parameter
of the model.

### Mean of the distribution

\[
T(F_i)=E(Y_i).
\]

### Variance

\[
T(F_i)=\operatorname{Var}(Y_i).
\]

### Quantile of order \(p\)

\[
T(F_i)=Q_i(p)=F_i^{-1}(p).
\]

### Exact mass at zero

\[
T(F_i)=P(Y_i=0).
\]

It is important to distinguish

\[
P(Y=0)
\]

from

\[
F(0)=P(Y\le 0).
\]

For a standard Normal continuous variable, for instance,

\[
F(0)=0.5,
\]

but

\[
P(Y=0)=0.
\]

The package uses the **exact point mass** for `prob_zero`.

---

## 3.3. Standardization over a target population

If \(T(F_{ig})\) is the estimand predicted for unit \(i\) under group \(g\),
a standardized mean can be written as

\[
\bar T_g
=
\frac{\sum_{i=1}^{N} w_i T(F_{ig})}
{\sum_{i=1}^{N}w_i}.
\]

The package allows three conceptual populations.

### `population = "observed"`

The observed covariates are preserved and the focal group is changed
counterfactually. This strategy corresponds to a form of **g-computation, or
standardization over the empirical distribution of the covariates**.

### `population = "balanced"`

A balanced population is built over the relevant combinations of the factors.

### `population = "reference"`

A reference grid is used, an approach especially related to the concept of
estimated marginal means (Lenth, 2016).

For this reason, two analyses may produce different results and both still be
correct: they may be estimating quantities for different target populations.

---

## 3.4. Contrasts on a scientific scale

`gamlssPosthoc` separates:

1. **who is compared**;
2. **how the effect is expressed**.

### Geometry

```{r illustrative-03, eval=FALSE}
contrast = "pairwise"
contrast = "reference"
contrast = "sequential"
contrast = "poly"
contrast = "opoly"
```

### Scale

```{r illustrative-04, eval=FALSE}
comparison = "difference"
comparison = "ratio"
comparison = "log_ratio"
comparison = "percent_change"
```

For two groups \(A\) and \(B\):

### Difference

\[
C_{\Delta}=T_A-T_B.
\]

### Ratio

\[
C_R=\frac{T_A}{T_B}.
\]

### Log-ratio

\[
C_{\log R}
=
\log\left(\frac{T_A}{T_B}\right).
\]

### Percent change

\[
C_{\%}
=
100
\left(
\frac{T_A}{T_B}-1
\right).
\]

This definition avoids ambiguous inferences produced only by the link scale.

---

## 3.5. Zero-adjusted distributions

Semicontinuous responses may contain a point mass at zero and a continuous
distribution for the positive values. This structure is common in
biological, environmental and agricultural applications (Stasinopoulos et
al., 2023).

Consider:

\[
P(Y=0)=h,
\]

and

\[
Y_+ \sim F_+,
\qquad Y_+>0.
\]

The complete distribution is then

\[
P(Y=0)=h,
\]

\[
f_Y(y)
=
(1-h)f_+(y),
\qquad y>0.
\]

### Mean

If

\[
m_+=E(Y_+),
\]

then

\[
\boxed{
E(Y)
=
(1-h)m_+
}
\]

### Variance

If

\[
v_+=\operatorname{Var}(Y_+),
\]

then

\[
\boxed{
\operatorname{Var}(Y)
=
(1-h)v_+
+
h(1-h)m_+^2
}
\]

### Quantile

\[
Q_Y(p)=0,
\qquad p\le h,
\]

and, for \(p>h\),

\[
\boxed{
Q_Y(p)
=
Q_+
\left(
\frac{p-h}{1-h}
\right)
}
\]

`gamlssPosthoc` applies this logic generically whenever the positive
distribution can be reconstructed by `gamlss.dist::GAMLSS()` and
`distributions3`.

---

## 3.6. A special case: Gamma and GAF

In the Gamma family `GA` of `gamlss.dist`,

\[
E(Y)=\mu,
\]

\[
\operatorname{Var}(Y)=\sigma^2\mu^2.
\]

In `GAF`, the exponent of the mean-variance relationship is flexible:

\[
E(Y)=\mu,
\]

\[
\boxed{
\operatorname{Var}(Y)
=
\sigma^2\mu^\nu
}
\]

(Enki et al., 2017; documentation of `gamlss.dist::GAF`).

When

\[
\nu=2,
\]

the mean-variance relationship coincides with the conventional Gamma.

The `GAF` documentation also stresses the possibly high correlation between
\(\sigma\) and \(\nu\), which is why inference on dispersion and power
parameters must be interpreted with care.

---

# 4. Conceptual architecture of the package

Version 0.2.0 uses three engines.

## 4.1. `emmeans`

Preferred when:

- the target is a simple distributional parameter;
- the selected model contains no smoother;
- the desired population is explicitly a reference grid;
- the contrast is compatible with the `emmeans` mechanism.

Using this tool preserves the advantages of the EMM and contrast
infrastructure developed by Lenth (2016).

---

## 4.2. `marginaleffects`

It can take greater responsibility when:

- the target is an ordinary distributional parameter;
- the analysis requires standardization by predictions;
- comparisons on different scales are wanted;
- a first-order derivative is wanted;
- the model API adequately supports the operation.

The `marginaleffects` paradigm treats predictions, slopes, comparisons and
other quantities as substantively interpretable transformations of the model
predictions (Arel-Bundock, Greifer & Heiss, 2024).

---

## 4.3. Distribution engine

This is the most general option when the estimand depends on:

- the complete distribution;
- multiple parameters;
- the mass at zero;
- smoothers;
- quantiles or variances;
- a custom function.

The central principle is:

\[
(\hat\theta_1,\ldots,\hat\theta_p)
\longrightarrow
\widehat F
\longrightarrow
T(\widehat F).
\]

This approach is coherent with the philosophy of distributional regression:
the predicted distribution is obtained first, and the scientific quantity of
interest is computed afterwards.

---

# 5. Installation and dependencies

A complete installation may require:

```{r illustrative-05, eval=FALSE}
install.packages(c(
  "gamlss",
  "gamlss.dist",
  "distributions3",
  "gamlss.inf",
  "emmeans",
  "marginaleffects",
  "future",
  "future.apply",
  "multcompView"
))
```

After installing the package:

```{r illustrative-06, eval=FALSE}
library(gamlssPosthoc)
library(gamlss)
library(gamlss.dist)
```

The additional dependencies are optional and are activated according to the
engine or functionality requested.

---

# 6. `gamlss_posthoc()`

## 6.1. Purpose

`gamlss_posthoc()` is the central function of the package.

It integrates:

- the choice of estimand;
- the definition of the target population;
- weighting;
- engine routing;
- computation of contrasts;
- the choice of scientific scale;
- uncertainty propagation.

### Main signature

```{r illustrative-07, eval=FALSE}
gamlss_posthoc(
  object,
  specs,
  by = NULL,
  estimand = c(
    "parameter", "mean", "variance",
    "quantile", "prob_zero", "custom"
  ),
  what = "mu",
  contrast = c(
    "none", "pairwise", "reference",
    "sequential", "poly", "opoly"
  ),
  comparison = c(
    "difference", "ratio",
    "log_ratio", "percent_change"
  ),
  adjust = "tukey",
  at = list(),
  population = NULL,
  weights = NULL,
  grid = NULL,
  engine = c(
    "auto", "emmeans",
    "marginaleffects", "distribution"
  ),
  uncertainty = c(
    "auto", "delta",
    "simulation", "bootstrap", "none"
  ),
  bootstrap = c("parametric", "case", "cluster"),
  B = 499L,
  cluster = NULL,
  ref = 1L,
  scores = NULL,
  degree = NULL,
  prob = 0.5,
  custom_fun = NULL,
  positive_dist_fun = NULL,
  positive_mean_fun = NULL,
  data = NULL,
  refit_fun = NULL,
  simulate_fun = NULL,
  parallel = FALSE,
  cores = 2L,
  seed = 1234,
  keep_draws = FALSE,
  ...
)
```

---

## 6.2. Structure of the returned object

The result has class `gamlss_posthoc` and contains, among others:

```{r illustrative-08, eval=FALSE}
result$estimates
result$contrasts
result$engine
result$estimand
result$estimand_info
result$population
result$weighting
result$uncertainty_method
```

When requested:

```{r illustrative-09, eval=FALSE}
result$draws
```

may contain bootstrap draws.

---

## 6.3. Worked example: a Gamma factorial experiment

In this example, three treatments are evaluated under two water regimes.

```{r factorial-gamma, eval=has_core}
library(gamlss)
library(gamlss.dist)
library(gamlssPosthoc)

set.seed(101)

dat <- expand.grid(
  treatment = factor(c("Control", "BioA", "BioB")),
  water = factor(c("100_ETc", "70_ETc")),
  block = factor(1:6),
  replicate = 1:4
)

eta <- with(
  dat,
  log(4.0) +
    0.18 * (treatment == "BioA") +
    0.32 * (treatment == "BioB") -
    0.25 * (water == "70_ETc") -
    0.08 * (treatment == "BioB" & water == "70_ETc")
)

mu <- exp(eta)

dat$biomass <- rGA(
  nrow(dat),
  mu = mu,
  sigma = 0.22
)

fit <- gamlss(
  biomass ~ treatment * water,
  sigma.formula = ~ water,
  family = GA,
  data = dat,
  trace = FALSE
)
```

### Marginal means of the response

Because `GA` has \(E(Y)=\mu\), the mean of the distribution coincides with the
`mu` parameter. Even so, using `estimand = "mean"` makes the scientific
intention explicit.

```{r marginal-means, eval=has_core}
res_mean <- gamlss_posthoc(
  fit,
  specs = "treatment",
  by = "water",
  estimand = "mean",
  population = "observed",
  contrast = "pairwise",
  comparison = "difference",
  uncertainty = "none",
  data = dat
)

res_mean
res_mean$estimates
res_mean$contrasts
```

### Ratio of means

```{r ratio-of-means, eval=has_core}
res_ratio <- gamlss_posthoc(
  fit,
  specs = "treatment",
  by = "water",
  estimand = "mean",
  population = "observed",
  contrast = "reference",
  comparison = "ratio",
  ref = 1,
  uncertainty = "none",
  data = dat
)

res_ratio$contrasts
```

### Percent change

```{r percent-change, eval=has_core}
res_pc <- gamlss_posthoc(
  fit,
  specs = "treatment",
  by = "water",
  estimand = "mean",
  population = "observed",
  contrast = "reference",
  comparison = "percent_change",
  ref = 1,
  uncertainty = "none",
  data = dat
)

res_pc$contrasts
```

An estimate of 25 for `percent_change` should be read as:

> the predicted marginal mean was approximately 25% higher than the reference
> category, under the declared target population.

---

## 6.4. Comparing `sigma`

The post-hoc analysis may also refer to another parameter of the
distribution.

```{r compare-sigma, eval=has_emm}
res_sigma <- gamlss_posthoc(
  fit,
  specs = "water",
  estimand = "parameter",
  what = "sigma",
  population = "reference",
  contrast = "pairwise",
  comparison = "difference",
  engine = "emmeans",
  uncertainty = "delta",
  data = dat
)

res_sigma
```

In this case, mean biomass is not being compared. The `sigma` parameter is.

---

## 6.5. Quantiles

To compare predictive medians:

```{r median, eval=has_dist}
res_median <- gamlss_posthoc(
  fit,
  specs = "treatment",
  by = "water",
  estimand = "quantile",
  prob = 0.50,
  population = "observed",
  contrast = "pairwise",
  comparison = "difference",
  engine = "distribution",
  uncertainty = "none",
  data = dat
)

res_median$estimates
```

For the 90th percentile:

```{r quantile-90, eval=has_dist}
res_q90 <- gamlss_posthoc(
  fit,
  specs = "treatment",
  estimand = "quantile",
  prob = 0.90,
  population = "observed",
  engine = "distribution",
  uncertainty = "none",
  data = dat
)
```

---

## 6.6. Variance

```{r variance, eval=has_dist}
res_var <- gamlss_posthoc(
  fit,
  specs = "treatment",
  by = "water",
  estimand = "variance",
  population = "observed",
  contrast = "pairwise",
  comparison = "ratio",
  uncertainty = "none",
  data = dat
)
```

This analysis can answer a question that differs from a comparison of means:

> do particular treatments change the heterogeneity of the response, even
> when the mean changes little?

---

## 6.7. Custom estimand

The interface allows user-defined derived functions.

For example, when the custom function can be computed from the distributional
predictions:

```{r illustrative-18, eval=FALSE}
res_custom <- gamlss_posthoc(
  fit,
  specs = "treatment",
  estimand = "custom",
  custom_fun = function(...) {
    # user function compatible with the interface
    # of the custom estimand
  },
  population = "observed",
  engine = "distribution",
  uncertainty = "none",
  data = dat
)
```

This facility is appropriate when the scientific question does not coincide
with the native estimands.

---

# 7. `gamlss_posthoc_plan()`

## 7.1. Purpose

`gamlss_posthoc_plan()` should be used **before** complex analyses.

It inspects:

- the model class;
- the family;
- the available parameters;
- the formulas;
- the links;
- the smoothers;
- the installed dependencies;
- engine eligibility;
- the recommended uncertainty;
- structural warnings.

### Signature

```{r illustrative-19, eval=FALSE}
gamlss_posthoc_plan(
  object,
  estimand = "parameter",
  what = "mu",
  contrast = "none",
  comparison = "difference",
  population = "observed",
  weights = NULL,
  uncertainty = "auto"
)
```

---

## 7.2. Worked example with `pb()`

```{r pb-data, eval=has_core}
library(gamlss)
library(gamlss.dist)
library(gamlssPosthoc)

set.seed(202)

n <- 180

dat <- data.frame(
  dose = runif(n, 0, 150),
  treatment = factor(
    sample(c("A", "B"), n, replace = TRUE)
  )
)

mu <- with(
  dat,
  exp(
    log(2.2) +
      0.010 * dose -
      0.00004 * dose^2 +
      0.18 * (treatment == "B")
  )
)

dat$biomass <- rGA(
  n,
  mu = mu,
  sigma = 0.25
)

fit_pb <- gamlss(
  biomass ~ treatment + pb(dose),
  sigma.formula = ~ treatment,
  family = GA,
  data = dat,
  trace = FALSE
)
```

Diagnostic for `mu`:

```{r plan-mu, eval=has_core}
plan_mu <- gamlss_posthoc_plan(
  fit_pb,
  estimand = "parameter",
  what = "mu",
  population = "reference",
  uncertainty = "auto"
)

print(plan_mu)
```

The plan is expected to identify the smoother in the `mu` submodel and to
avoid recommending `emmeans` for that component.

Diagnostic for `sigma`:

```{r plan-sigma, eval=has_core}
plan_sigma <- gamlss_posthoc_plan(
  fit_pb,
  estimand = "parameter",
  what = "sigma",
  population = "reference",
  uncertainty = "auto"
)

print(plan_sigma)
```

Because `sigma.formula` contains no `pb()`, eligibility may differ.

---

## 7.3. Why the plan matters

Automatic engine selection should not be treated as a black box. The plan
lets the analyst see in advance:

\[
\text{requested estimand}
+
\text{model structure}
+
\text{population}
+
\text{uncertainty}
\Rightarrow
\text{recommended strategy}.
\]

This makes the analysis auditable.

---

# 8. `gamlss_distribution_summary()`

## 8.1. Purpose

`gamlss_distribution_summary()` returns plug-in summaries of the predicted
distribution:

- mean;
- variance;
- standard deviation;
- exact mass at zero;
- requested quantiles.

### Signature

```{r illustrative-23, eval=FALSE}
gamlss_distribution_summary(
  object,
  newdata,
  probs = c(0.025, 0.5, 0.975),
  data = NULL,
  positive_dist_fun = NULL
)
```

---

## 8.2. Worked example: ordinary Gamma

```{r gamma-summary, eval=has_dist}
set.seed(303)

dat <- data.frame(
  N = runif(100, 40, 160)
)

dat$yield <- rGA(
  100,
  mu = exp(2.0 + 0.004 * dat$N),
  sigma = 0.18
)

fit_ga <- gamlss(
  yield ~ N,
  family = GA,
  data = dat,
  trace = FALSE
)

newd <- data.frame(
  N = c(50, 100, 150)
)

sum_ga <- gamlss_distribution_summary(
  fit_ga,
  newdata = newd,
  probs = c(0.05, 0.50, 0.95),
  data = dat
)

sum_ga
```

For each value of `N`, the function returns a description of the predicted
distribution, not merely a single parameter.

---

## 8.3. Worked example: zero-adjusted Generalized Gamma

This is a case where the advantage of the distribution engine becomes
evident.

```{r zagg-data, eval=has_zadj}
library(gamlss.inf)

set.seed(304)

dat <- data.frame(
  treatment = factor(
    rep(c("Control", "Inoculated"), each = 100)
  )
)

mu <- ifelse(
  dat$treatment == "Control",
  1.8,
  3.0
)

h <- ifelse(
  dat$treatment == "Control",
  0.45,
  0.20
)

positive_part <- rGG(
  nrow(dat),
  mu = mu,
  sigma = 0.35,
  nu = 0.7
)

dat$nodules <- ifelse(
  rbinom(nrow(dat), 1, h) == 1,
  0,
  positive_part
)

fit_zagg <- gamlssZadj(
  y = nodules,
  mu.formula = ~ treatment,
  sigma.formula = ~ 1,
  nu.formula = ~ 1,
  xi0.formula = ~ treatment,
  family = GG,
  data = dat,
  trace = FALSE
)
```

Creating predictions:

```{r zagg-summary, eval=has_zadj}
newd <- data.frame(
  treatment = factor(
    c("Control", "Inoculated"),
    levels = levels(dat$treatment)
  )
)

sum_zagg <- gamlss_distribution_summary(
  fit_zagg,
  newdata = newd,
  probs = c(0.10, 0.50, 0.90),
  data = dat
)

sum_zagg
```

The function automatically combines

\[
P(Y=0)=\xi_0
\]

with the positive `GG` distribution.

---

## 8.4. Agronomic interpretation

If the inoculated treatment reduces the mass at zero and increases the
positive part, the marginal mean may change through two mechanisms:

\[
\text{treatment}
\rightarrow
P(Y>0)
\]

and

\[
\text{treatment}
\rightarrow
E(Y\mid Y>0).
\]

This decomposition is biologically more informative than a simple arithmetic
mean of the observed data.

---

# 9. `gamlss_trend()`

## 9.1. Purpose

`gamlss_trend()` was developed for quantitative predictors.

It allows:

```{r illustrative-27, eval=FALSE}
method = "curve"
method = "derivative"
method = "turning_points"
method = "optimum"
```

In addition:

```{r illustrative-28, eval=FALSE}
derivative_order = 1
derivative_order = 2
```

and

```{r illustrative-29, eval=FALSE}
optimum = "maximum"
optimum = "minimum"
```

### Simplified signature

```{r illustrative-30, eval=FALSE}
gamlss_trend(
  object,
  x,
  by = NULL,
  at_x = NULL,
  n = 100L,
  method = c(
    "curve", "derivative",
    "turning_points", "optimum"
  ),
  derivative_order = 1L,
  optimum = c("maximum", "minimum"),
  band = c("pointwise", "simultaneous"),
  estimand = "mean",
  what = "mu",
  engine = c(
    "auto", "marginaleffects", "distribution"
  ),
  population = NULL,
  weights = NULL,
  data = NULL,
  uncertainty = c(
    "auto", "delta",
    "simulation", "bootstrap", "none"
  ),
  bootstrap = "case",
  B = 499L,
  ...
)
```

---

## 9.2. Numerical derivatives

For a curve

\[
m(x)=E(Y\mid x),
\]

the first derivative is

\[
m'(x)
=
\frac{dm(x)}{dx}.
\]

On an interior grid, the central-difference approximation used conceptually
is

\[
m'(x_i)
\approx
\frac{
m(x_{i+1})-m(x_{i-1})
}{
x_{i+1}-x_{i-1}
}.
\]

The second derivative is

\[
m''(x)
=
\frac{d^2m(x)}{dx^2}.
\]

It provides information about curvature.

---

## 9.3. Worked example: quadratic dose response

```{r dose-data, eval=has_core}
set.seed(401)

n <- 220

dat <- data.frame(
  dose = runif(n, 0, 180)
)

eta <- with(
  dat,
  log(2.0) +
    0.014 * dose -
    0.000055 * dose^2
)

dat$biomass <- rGA(
  n,
  mu = exp(eta),
  sigma = 0.20
)

fit_dose <- gamlss(
  biomass ~ dose + I(dose^2),
  family = GA,
  data = dat,
  trace = FALSE
)

grid_dose <- seq(0, 180, length.out = 181)
```

### Curve

```{r trend-curve, eval=has_core}
trend_curve <- gamlss_trend(
  fit_dose,
  x = "dose",
  at_x = grid_dose,
  method = "curve",
  estimand = "mean",
  uncertainty = "none",
  data = dat
)

plot(trend_curve)
```

### First derivative

```{r derivative1, eval=has_core}
d1 <- gamlss_trend(
  fit_dose,
  x = "dose",
  at_x = grid_dose,
  method = "derivative",
  derivative_order = 1,
  estimand = "mean",
  uncertainty = "none",
  data = dat
)

plot(d1)
```

Interpretation:

- \(m'(x)>0\): the response still increases with dose;
- \(m'(x)<0\): the response decreases;
- \(m'(x)\approx0\): a stationary region.

### Second derivative

```{r derivative2, eval=has_core}
d2 <- gamlss_trend(
  fit_dose,
  x = "dose",
  at_x = grid_dose,
  method = "derivative",
  derivative_order = 2,
  estimand = "mean",
  uncertainty = "none",
  data = dat
)
```

### Stationary points

```{r turning-points, eval=has_core}
tp <- gamlss_trend(
  fit_dose,
  x = "dose",
  at_x = grid_dose,
  method = "turning_points",
  estimand = "mean",
  uncertainty = "none",
  data = dat
)

tp$special_points
```

### Maximum

```{r maximum, eval=has_core}
opt <- gamlss_trend(
  fit_dose,
  x = "dose",
  at_x = grid_dose,
  method = "optimum",
  optimum = "maximum",
  estimand = "mean",
  uncertainty = "none",
  data = dat
)

opt$special_points
```

The package performs a local quadratic refinement when the extremum occurs in
the interior of the grid and the local curvature is compatible with a maximum
or a minimum.

---

## 9.4. Example with `pb()`

```{r curve-pb, eval=has_core}
fit_smooth <- gamlss(
  biomass ~ pb(dose),
  family = GA,
  data = dat,
  trace = FALSE
)

smooth_curve <- gamlss_trend(
  fit_smooth,
  x = "dose",
  at_x = grid_dose,
  method = "curve",
  estimand = "mean",
  engine = "distribution",
  uncertainty = "none",
  data = dat
)

plot(smooth_curve)
```

This is a central case for the package, because inference is based on the
predicted curve, without requiring `emmeans` to interpret the smoother.

---

# 10. `gamlss_poly_compare()`

## 10.1. Purpose

`gamlss_poly_compare()` compares **already fitted** and nested polynomial
models.

The function was deliberately kept separate from polynomial contrasts.

### Polynomial regression

Dose is quantitative:

\[
\eta(x)
=
\beta_0+
\beta_1x+
\beta_2x^2+\cdots.
\]

### Polynomial contrast

The levels are treated as an ordered set and linear combinations of the means
are built.

The two analyses answer different questions.

---

## 10.2. GAIC

The function computes

\[
GAIC
=
GD+K\,df,
\]

where:

- \(GD\) is the global deviance;
- \(df\) is the effective number of degrees of freedom;
- \(K\) is the penalty.

When

\[
K=2,
\]

the form corresponding to AIC is obtained.

A choice such as

\[
K=\log(n)
\]

produces a penalty analogous to that of BIC.

---

## 10.3. Likelihood ratio

For nested models,

\[
LR
=
GD_{\text{simple}}
-
GD_{\text{complex}}.
\]

If the difference in degrees of freedom is

\[
\Delta df,
\]

the approximation used is

\[
LR
\sim
\chi^2_{\Delta df}.
\]

---

## 10.4. Worked example

```{r poly-compare, eval=has_core}
set.seed(501)

dat <- data.frame(
  dose = rep(seq(0, 160, by = 20), each = 15)
)

mu <- with(
  dat,
  exp(
    log(3.0) +
      0.010 * dose -
      0.000035 * dose^2
  )
)

dat$yield <- rGA(
  nrow(dat),
  mu = mu,
  sigma = 0.18
)

m1 <- gamlss(
  yield ~ dose,
  family = GA,
  data = dat,
  trace = FALSE
)

m2 <- gamlss(
  yield ~ dose + I(dose^2),
  family = GA,
  data = dat,
  trace = FALSE
)

m3 <- gamlss(
  yield ~ dose + I(dose^2) + I(dose^3),
  family = GA,
  data = dat,
  trace = FALSE
)

gaic_table <- gamlss_poly_compare(
  linear = m1,
  quadratic = m2,
  cubic = m3,
  k = 2
)

gaic_table
```

Selection should not be based on the \(p\)-value alone. The following must
also be considered:

- agronomic plausibility;
- the shape of the curve;
- model diagnostics;
- the gain in fit;
- the stability of the optimum;
- extrapolation.

---

# 11. `gamlss_cld()`

## 11.1. Purpose

`gamlss_cld()` builds a **compact letter display (CLD)** from pairwise
comparisons.

The letter representation is a compact way of summarising a matrix of
multiple comparisons (Piepho, 2004).

### Important

The letters are a **presentation layer**, not the main inferential result.

It is advisable to present, as a priority:

- estimates;
- confidence intervals;
- contrasts;
- the scale of the effect;
- adjusted \(p\)-values.

---

## 11.2. Worked example

```{r cld-example, eval=has_cld}
set.seed(601)

dat <- data.frame(
  treatment = factor(
    rep(c("T0", "T1", "T2", "T3"), each = 35)
  )
)

mu <- c(
  T0 = 3.0,
  T1 = 3.2,
  T2 = 4.0,
  T3 = 4.1
)

dat$y <- rGA(
  nrow(dat),
  mu = mu[dat$treatment],
  sigma = 0.20
)

fit <- gamlss(
  y ~ treatment,
  family = GA,
  data = dat,
  trace = FALSE
)

res <- gamlss_posthoc(
  fit,
  specs = "treatment",
  estimand = "parameter",
  what = "mu",
  population = "reference",
  contrast = "pairwise",
  comparison = "difference",
  engine = "emmeans",
  uncertainty = "delta",
  data = dat
)

res$contrasts

cld_out <- gamlss_cld(
  res,
  alpha = 0.05
)

cld_out
```

Two treatments sharing a letter means that the procedure used did not detect
a difference between them at the stated level. This **does not prove
equivalence** between the treatments.

---

# 12. `gamlss_engine_info()`

## 12.1. Purpose

`gamlss_engine_info()` is a compact wrapper kept for backward compatibility.

New analyses should prefer:

```{r illustrative-40, eval=FALSE}
gamlss_posthoc_plan()
```

---

## 12.2. Worked example

```{r engine-info, eval=has_core}
set.seed(701)

dat <- data.frame(
  dose = runif(120, 0, 100)
)

dat$y <- rGA(
  120,
  mu = exp(1 + 0.01 * dat$dose),
  sigma = 0.20
)

fit <- gamlss(
  y ~ pb(dose),
  family = GA,
  data = dat,
  trace = FALSE
)

gamlss_engine_info(
  fit,
  estimand = "parameter",
  what = "mu",
  population = "reference"
)
```

For a fuller inspection:

```{r engine-plan, eval=has_core}
gamlss_posthoc_plan(
  fit,
  estimand = "parameter",
  what = "mu",
  population = "reference"
)
```

---

# 13. Recommended analysis workflows

## 13.1. Qualitative treatment, simple parametric model

```text
fit GAMLSS
      |
      v
diagnose the model
      |
      v
gamlss_posthoc_plan()
      |
      v
define the estimand
      |
      v
gamlss_posthoc()
      |
      v
estimates + CI + contrasts
      |
      v
CLD only if necessary
```

---

## 13.2. Quantitative dose

```text
fit the curve
      |
      v
compare candidate models
      |
      v
gamlss_poly_compare()
      |
      v
gamlss_trend(method = "curve")
      |
      v
gamlss_trend(method = "derivative")
      |
      v
turning point / optimum
      |
      v
agronomic interpretation
```

---

## 13.3. `pb()` smoother

```text
gamlss(... ~ pb(x))
      |
      v
gamlss_posthoc_plan()
      |
      v
distribution engine
      |
      v
predicted curve
      |
      v
derivatives / optimum
      |
      v
bootstrap if a CI is needed
```

---

## 13.4. Zero-adjusted

```text
gamlssZadj
   |
   v
predict xi0
   +
predict the positive distribution
   |
   v
rebuild F(Y|X)
   |
   v
E(Y), Var(Y), Qp, P(Y=0)
   |
   v
standardize
   |
   v
contrast
```

---

# 14. Zero-adjusted models

## 14.1. Why comparing `mu` alone is not enough

Consider two treatments.

Treatment A:

\[
h_A=0.50,
\qquad
E(Y_+\mid A)=4.
\]

Treatment B:

\[
h_B=0.10,
\qquad
E(Y_+\mid B)=3.
\]

Comparing the positive part alone:

\[
4>3.
\]

But the marginal means are

\[
E(Y\mid A)
=
(1-0.50)4
=
2,
\]

and

\[
E(Y\mid B)
=
(1-0.10)3
=
2.7.
\]

The conclusion is reversed.

This example shows why defining the estimand is central.

---

## 14.2. Example with nodule mass

A nodule-mass response may contain:

- plants without nodulation: \(Y=0\);
- nodulated plants: \(Y>0\).

The treatment effect can be decomposed into:

\[
\Delta P(Y>0)
\]

and

\[
\Delta E(Y\mid Y>0).
\]

The marginal mean incorporates both mechanisms:

\[
E(Y)
=
P(Y>0)
E(Y\mid Y>0).
\]

This kind of interpretation is often more useful biologically than a single
comparison of parameters.

---

# 15. Comparisons, multiplicity and interpretation

## 15.1. Multiplicity adjustment

In multiple comparisons, the probability of extreme results grows with the
number of tests.

`gamlssPosthoc` allows adjustment methods compatible with the infrastructure
in use to be passed through.

Examples:

```{r illustrative-47, eval=FALSE}
adjust = "holm"
adjust = "bonferroni"
adjust = "BH"
```

When the engine is `emmeans`, adjustments specific to that infrastructure can
be used where applicable.

In the bootstrap engine, methods from `stats::p.adjust()` are used. If a
request is not available in that context, the package avoids presenting a
different adjustment as though it were the requested method.

---

## 15.2. Avoiding dichotomous interpretation

A conclusion should not be reduced to:

> "A differs from B because \(p<0.05\)."

It is preferable to report

\[
\widehat{\Delta},
\]

the confidence interval,

\[
[\Delta_L,\Delta_U],
\]

and the agronomic relevance of the effect.

For example:

> Treatment B showed an estimated marginal mean 18% higher than the control,
> with a confidence interval compatible with increases between approximately
> X% and Y%.

---

## 15.3. Polynomial contrasts

When the levels represent doses treated as an ordered factor, one may use:

```{r illustrative-48, eval=FALSE}
contrast = "poly"
```

For unequally spaced levels:

```{r illustrative-49, eval=FALSE}
contrast = "opoly"
scores = c(0, 25, 50, 100)
```

This is distinct from fitting

```{r illustrative-50, eval=FALSE}
y ~ dose + I(dose^2)
```

as a quantitative regression.

---

# 16. Quantitative regression, derivatives and optima

## 16.1. Why derivatives are useful

In experiments with dose, concentration, salinity, time or temperature, the
scientific question is often:

> Over which range does the response still increase?

The first derivative answers directly:

\[
m'(x)>0
\Rightarrow
\text{increasing response},
\]

\[
m'(x)<0
\Rightarrow
\text{decreasing response}.
\]

---

## 16.2. Point of maximum

For a quadratic regression on the appropriate scale,

\[
\eta(x)
=
\beta_0+\beta_1x+\beta_2x^2,
\]

the stationary point of the linear predictor is

\[
x^*
=
-\frac{\beta_1}{2\beta_2}.
\]

In GAMLSS, however:

- the link may be non-linear;
- the estimand may not be the linear predictor;
- the mean may depend on several parameters;
- the curve may be a smoother.

For this reason `gamlss_trend()` works preferentially on the **curve of the
declared estimand**, and not on an assumed closed-form formula.

---

## 16.3. A statistical maximum is not necessarily a recommended dose

The predicted maximum of an agronomic variable should not automatically be
called an "optimal dose".

A recommendation may require:

- economic analysis;
- toxicity;
- environmental risk;
- operational constraints;
- stability across environments;
- other outcomes.

Thus `method = "optimum"` locates a mathematical extremum of the estimand.
Its scientific interpretation depends on the context.

---

# 17. Uncertainty: delta, simulation and bootstrap

## 17.1. Delta method

For a differentiable function

\[
g(\boldsymbol{\beta}),
\]

the variance can be approximated by

\[
\operatorname{Var}
\left[
g(\hat{\boldsymbol{\beta}})
\right]
\approx
\nabla g(\hat{\boldsymbol{\beta}})^{\mathsf T}
\widehat{\operatorname{Var}}(\hat{\boldsymbol{\beta}})
\nabla g(\hat{\boldsymbol{\beta}}).
\]

The method is efficient when a suitable covariance matrix exists and the
transformation is regular.

---

## 17.2. Simulation from the coefficient uncertainty

When a multivariate approximation of the coefficients is appropriate,

\[
\boldsymbol{\beta}^{(b)}
\sim
N
\left(
\hat{\boldsymbol{\beta}},
\widehat{\mathbf V}
\right),
\]

and the estimand is recomputed for each draw.

This strategy can be used by engines that provide the necessary
infrastructure.

---

## 17.3. Refit bootstrap

The bootstrap is the safest generic option when the estimand depends on:

- several parameters;
- a zero-adjusted transformation;
- a smoother;
- a complex non-linear function;
- a curve, a derivative or an optimum.

For \(b=1,\ldots,B\):

1. generate or resample data;
2. refit the model;
3. recompute the estimand;
4. store \(T^{(b)}\).

The standard error can be estimated by

\[
SE_{\text{boot}}
=
\sqrt{
\frac{1}{B-1}
\sum_{b=1}^{B}
\left(
T^{(b)}-\bar T
\right)^2
}.
\]

A 95% percentile interval uses

\[
Q_{0.025}
\{T^{(1)},\ldots,T^{(B)}\}
\]

and

\[
Q_{0.975}
\{T^{(1)},\ldots,T^{(B)}\}.
\]

---

## 17.4. Types of bootstrap

### Parametric

```{r illustrative-51, eval=FALSE}
bootstrap = "parametric"
```

Data are generated from the fitted model.

### Case

```{r illustrative-52, eval=FALSE}
bootstrap = "case"
```

Observations are resampled.

### Cluster

```{r illustrative-53, eval=FALSE}
bootstrap = "cluster"
cluster = "block"
```

Whole clusters are resampled.

For dependent experimental data, the resampling unit must respect the
experimental unit and the design.

---

## 17.5. Simultaneous bands

For a curve evaluated at several points, pointwise intervals do not control
the simultaneous uncertainty over the whole trajectory.

With bootstrap draws, the package can obtain an approximate maximum
standardized deviation statistic:

\[
M^{(b)}
=
\max_j
\left|
\frac{
T_j^{(b)}-\hat T_j
}{
\widehat{SE}_j
}
\right|.
\]

If

\[
c_{0.95}
=
Q_{0.95}(M),
\]

the simultaneous band is built approximately as

\[
\hat T_j
\pm
c_{0.95}\widehat{SE}_j.
\]

---

# 18. Good practice and limitations

## 18.1. Post-hoc analysis does not fix an inadequate model

The model must be evaluated before the comparisons.

The following should be considered:

- residuals;
- worm plots;
- adequacy of the family;
- influence;
- convergence;
- dependence structure;
- plausibility of the terms;
- quality of the predictions.

The GAMLSS tradition uses randomized quantile residuals for the diagnosis of
distributional models (Dunn & Smyth, 1996).

---

## 18.2. Do not interpret `mu` without checking the family

Before writing

> "adjusted mean of `mu`",

check whether

\[
\mu=E(Y)
\]

holds in the family being used.

If that equality does not hold, prefer

```{r illustrative-54, eval=FALSE}
estimand = "mean"
```

---

## 18.3. An observed zero needs a scientific interpretation

In a zero-adjusted distribution, the model separates zero from the positive
part.

But an observed zero may represent:

- biological absence;
- a value below the detection limit;
- censoring;
- rounding;
- measurement failure.

The zero-mass component must be coherent with the observational mechanism.

---

## 18.4. Avoid extrapolation

In quantitative regressions, do not interpret

- derivatives;
- turning points;
- maxima

outside the range of data that supports the fit.

---

## 18.5. `emmeans` should not be used as a universal engine

`emmeans` remains extremely useful in its natural domain, but the documented
support for `gamlss` is partial and does not cover the selected submodel when
smoothers such as `pb()` are present.

The package avoids forcing that route.

---

## 18.6. `marginaleffects` is used with a capability test

Support for a class by a generic package does not imply that every
combination of

- family;
- parameter;
- smoother;
- prediction function

will be valid.

For this reason `gamlssPosthoc` tests the operation at run time and, under
automatic routing, may fall back to the distribution engine.

---

## 18.7. The bootstrap can fail

A high rate of refits that do not converge is information about the stability
of the model.

The package should not silently discard those replicates without warning the
user.

---

# 19. Summary of the functions

| Function | Main purpose | Typical use |
|---|---|---|
| `gamlss_posthoc()` | General post-model inference | means, parameters, variances, quantiles, contrasts |
| `gamlss_posthoc_plan()` | Diagnosis and routing | deciding engine and uncertainty before the analysis |
| `gamlss_distribution_summary()` | Summarise the predicted distribution | mean, variance, quantiles, mass at zero |
| `gamlss_trend()` | Quantitative regression | curve, derivatives, turning points, maximum/minimum |
| `gamlss_poly_compare()` | Compare nested polynomial regressions | linear vs quadratic vs cubic |
| `gamlss_cld()` | Compact letter display | compact summary of pairwise comparisons |
| `gamlss_engine_info()` | Compact legacy diagnostic | compatibility with older workflows |

---

# 20. References

Arel-Bundock, V., Greifer, N., & Heiss, A. (2024). How to interpret statistical models using `marginaleffects` for R and Python. *Journal of Statistical Software, 111*(9). https://doi.org/10.18637/jss.v111.i09

Dunn, P. K., & Smyth, G. K. (1996). Randomized quantile residuals. *Journal of Computational and Graphical Statistics, 5*(3), 236-244. https://doi.org/10.1080/10618600.1996.10474708

Enki, D. G., Noufaily, A., Farrington, P., Garthwaite, P., Andrews, N., & Charlett, A. (2017). Taylor's power law and the statistical modelling of infectious disease surveillance data. *Journal of the Royal Statistical Society: Series A (Statistics in Society), 180*(1), 45-72. https://doi.org/10.1111/rssa.12181

Lenth, R. V. (2016). Least-squares means: The R package `lsmeans`. *Journal of Statistical Software, 69*(1), 1-33. https://doi.org/10.18637/jss.v069.i01

Piepho, H.-P. (2004). An algorithm for a letter-based representation of all-pairwise comparisons. *Journal of Computational and Graphical Statistics, 13*(2), 456-466. https://doi.org/10.1198/1061860043515

Rigby, R. A., & Stasinopoulos, D. M. (2005). Generalized additive models for location, scale and shape. *Journal of the Royal Statistical Society: Series C (Applied Statistics), 54*(3), 507-554. https://doi.org/10.1111/j.1467-9876.2005.00510.x

Rigby, R. A., Stasinopoulos, D. M., Heller, G. Z., & De Bastiani, F. (2019). *Distributions for Modeling Location, Scale, and Shape: Using GAMLSS in R*. Chapman & Hall/CRC. https://doi.org/10.1201/9780429298547

Stasinopoulos, D. M., & Rigby, R. A. (2007). Generalized additive models for location scale and shape (GAMLSS) in R. *Journal of Statistical Software, 23*(7), 1-46. https://doi.org/10.18637/jss.v023.i07

Stasinopoulos, D. M., Rigby, R. A., & De Bastiani, F. (2018). GAMLSS: A distributional regression approach. *Statistical Modelling, 18*(3-4), 248-273. https://doi.org/10.1177/1471082X18759144

Stasinopoulos, D. M., Rigby, R. A., Heller, G. Z., Voudouris, V., & De Bastiani, F. (2017). *Flexible Regression and Smoothing: Using GAMLSS in R*. Chapman & Hall/CRC. https://doi.org/10.1201/b21973

Stasinopoulos, D. M., Rigby, R. A., Heller, G. Z., & De Bastiani, F. (2023). P-splines and GAMLSS: A powerful combination, with an application to zero-adjusted distributions. *Statistical Modelling, 23*(5-6), 510-524. https://doi.org/10.1177/1471082X231176635

Umlauf, N., Klein, N., & Zeileis, A. (2018). BAMLSS: Bayesian additive models for location, scale and shape (and beyond). *Journal of Computational and Graphical Statistics, 27*(3), 612-627. https://doi.org/10.1080/10618600.2017.1407325

Umlauf, N., Klein, N., Simon, T., & Zeileis, A. (2021). `bamlss`: A Lego toolbox for flexible Bayesian regression (and beyond). *Journal of Statistical Software, 100*(4), 1-53. https://doi.org/10.18637/jss.v100.i04

---

## Software documentation consulted

**`emmeans`**
Lenth, R. V., & Piaskowski, J. *Models supported by emmeans*.
https://rvlenth.github.io/emmeans/articles/models.html

**`gamlss`**
Generalized Additive Models for Location Scale and Shape.
https://search.r-project.org/CRAN/refmans/gamlss/html/gamlss.html

**`gamlss.dist::GAMLSS()`**
Create a GAMLSS Distribution.
https://search.r-project.org/CRAN/refmans/gamlss.dist/html/GAMLSS.html

**`gamlss.dist::GA()`**
Gamma distribution for fitting a GAMLSS.
https://search.r-project.org/CRAN/refmans/gamlss.dist/html/GA.html

**`gamlss.dist::GAF()`**
Gamma family with flexible mean-variance power.
https://search.r-project.org/CRAN/refmans/gamlss.dist/html/GAF.html

**`gamlss.inf::gamlssZadj()`**
Fitting positive real-line response variables with zeros.
https://rdrr.io/cran/gamlss.inf/man/gamlssZadj.html

**`marginaleffects`**
Predictions, comparisons, slopes, marginal means, and hypothesis tests.
https://marginaleffects.com/

---

# Appendix A. Central methodological message

The main idea of `gamlssPosthoc` can be summarised in one sentence:

> **In distributional regression, the post-model analysis must be defined by
> the scientific estimand and by the target population, and not by the
> automatic availability of an "adjusted means" function.**

In mathematical terms:

\[
\boxed{
\widehat F(Y\mid X)
\rightarrow
T\{\widehat F(Y\mid X)\}
\rightarrow
\operatorname{Marginalization}
\rightarrow
\operatorname{Contrast}
\rightarrow
\operatorname{Uncertainty}
}
\]

This organisation makes it possible to apply the same logic to parameters,
means, variances, quantiles, point probabilities, zero-adjusted responses,
dose curves and other derived estimands.

For computationally heavy bootstrap examples, reduced values of `B` are used
in this vignette; larger values are recommended for real scientific analysis.

---

# Appendix B. S3 printing and plotting methods

Although they are not the main analytical API, the package registers S3
methods that make the results easier to inspect.

## `print.gamlss_posthoc()`

It is called automatically by

```{r illustrative-59, eval=FALSE}
res <- gamlss_posthoc(...)
print(res)
```

or simply

```{r illustrative-60, eval=FALSE}
res
```

The method explicitly displays:

- the engine used;
- the estimand;
- the recorded mathematical definition;
- the target population;
- the weighting;
- the uncertainty layer;
- the scale of the comparison;
- the estimates;
- the contrasts, when they exist.

This output was deliberately designed to prevent the result from being
interpreted without knowledge of the estimand.

## `plot.gamlss_posthoc()`

```{r illustrative-61, eval=FALSE}
plot(res)
```

Produces a basic visualisation of the estimates by group. When `lower.CL` and
`upper.CL` are available, the intervals are added to the plot.

The method is deliberately simple. For publication figures, it is advisable
to convert

```{r illustrative-62, eval=FALSE}
res$estimates
```

into a `data.frame` and build the figure with the graphical tool of the
user's choice.

## `print.gamlss_posthoc_plan()`

```{r illustrative-63, eval=FALSE}
plan <- gamlss_posthoc_plan(...)
print(plan)
```

Shows:

- an overview of the model;
- the distributional parameters;
- the capability of each parameter;
- the availability of the engines;
- the recommendation;
- the warnings.

## `print.gamlss_trend()`

```{r illustrative-64, eval=FALSE}
tr <- gamlss_trend(...)
print(tr)
```

Displays the quantitative method used, the predictor and the table of
results. If maxima, minima or stationary points exist, they are shown in a
separate section.

## `plot.gamlss_trend()`

```{r illustrative-65, eval=FALSE}
plot(tr)
```

For `method = "curve"`, it draws the curve of the estimand. For derivatives,
it draws the corresponding derivative. Special points, when present, are
overlaid.

These methods are inspection tools and do not replace the preparation of
complete scientific graphics.
