---
title: "Missing Data. Multiple Imputation, GAMLSS and Post-Hoc Inference"
author: "Walter Esfrain Pereira"
output: rmarkdown::html_vignette
bibliography: references.bib
link-citations: true
vignette: >
  %\VignetteIndexEntry{Missing Data. Multiple Imputation, GAMLSS and Post-Hoc Inference}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 6, fig.height = 3.6)
library(gamlssPosthoc)
has_core <- requireNamespace("gamlss", quietly = TRUE) &&
  requireNamespace("gamlss.dist", quietly = TRUE)
has_dist <- has_core && requireNamespace("distributions3", quietly = TRUE)
has_gg   <- has_core && requireNamespace("ggplot2", quietly = TRUE)
has_mice <- has_core && requireNamespace("mice", quietly = TRUE)
if (has_core) suppressPackageStartupMessages(library(gamlss.dist))
```

# 1. Principle

`gamlssPosthoc` does not implement imputation, and it should not. Imputation is
a modelling problem of its own, and delegating it to a specialized tool such as
`mice` [@vanbuuren2011] keeps the two model specifications visible and auditable.

The rule that organizes everything below is simple: **the same model and the same
estimand must be computed in every imputed data set, and only then combined.**
Combining first and estimating afterwards, or stacking the imputed data sets and
treating them as independent observations, produces standard errors that are too
small by roughly a factor of \(\sqrt{M}\).

# 2. Rubin's rules

Let \(\hat{Q}_m\) and \(U_m\) be the estimate and its within-imputation variance
in imputation \(m = 1, \ldots, M\). The pooled quantities are [@rubin1987]

\[
\bar{Q} = \frac{1}{M}\sum_{m=1}^{M} \hat{Q}_m,
\qquad
\bar{U} = \frac{1}{M}\sum_{m=1}^{M} U_m,
\tag{1}
\]

\[
B = \frac{1}{M-1}\sum_{m=1}^{M}\left(\hat{Q}_m - \bar{Q}\right)^2,
\qquad
T = \bar{U} + \left(1 + \frac{1}{M}\right) B .
\tag{2}
\]

The total variance \(T\) has two parts: the ordinary sampling variance
\(\bar{U}\), and the extra variance \(B\) caused by not knowing the missing
values. The factor \((1 + 1/M)\) is the finite-\(M\) correction.

## 2.1 How many imputations are enough

The diagnostic that answers this is the **fraction of missing information**,

\[
\gamma = \frac{\left(1 + M^{-1}\right) B}{T},
\tag{3}
\]

which is the share of the total variance attributable to missingness. A common
working rule is to take \(M\) at least equal to \(100\gamma\) rounded up, so a
data set with \(\gamma \approx 0.3\) needs \(M \approx 30\), not the \(M = 5\)
that older texts suggested. Report \(\gamma\); it tells the reader how much of
the uncertainty is imputation and how much is sampling.

## 2.2 Degrees of freedom in small samples

The classical pooled degrees of freedom,

\[
\nu_{\text{old}} = (M - 1)\left(1 + \frac{\bar{U}}{\left(1 + M^{-1}\right)B}\right)^{2},
\tag{4}
\]

can exceed the complete-data degrees of freedom, which is impossible. The
Barnard and Rubin correction [@barnard1999] repairs this. With \(\nu_{\text{com}}\)
the complete-data degrees of freedom,

\[
\hat\nu_{\text{obs}} = \frac{\nu_{\text{com}} + 1}{\nu_{\text{com}} + 3}\,
\nu_{\text{com}}\,(1 - \gamma),
\qquad
\nu_{\text{adj}} = \left(\frac{1}{\nu_{\text{old}}} +
\frac{1}{\hat\nu_{\text{obs}}}\right)^{-1}.
\tag{5}
\]

Use equation (5) whenever the analysis sample is small, which in agronomic field
trials is the normal case.

# 3. Executable workflow

The block below runs the complete cycle: generate missingness, impute, fit a
GAMLSS per imputation, compute the same post-hoc estimand, and pool. `M` is kept
small so the vignette builds quickly; use `M` guided by equation (3) in real
work.

```{r data, eval=has_mice}
set.seed(6001)
n <- 120
d <- data.frame(
  trt = factor(rep(c("Control", "Bio"), each = n / 2), levels = c("Control", "Bio")),
  x   = stats::rnorm(n, 50, 10)
)
d$y <- gamlss.dist::rGA(n,
  mu = exp(0.6 + 0.25 * (d$trt == "Bio") + 0.010 * (d$x - 50)),
  sigma = 0.30)

# MAR: the covariate is missing more often in the treated group
p_mis <- ifelse(d$trt == "Bio", 0.30, 0.10)
d$x[stats::rbinom(n, 1, p_mis) == 1] <- NA
sum(is.na(d$x))
```

```{r impute, eval=has_mice}
M <- 5L
imp <- mice::mice(d, m = M, printFlag = FALSE, seed = 6002)
imp$method
```

```{r analyse, eval=has_mice && has_dist}
one_fit <- function(m) {
  dm <- mice::complete(imp, m)
  fit <- gamlss::gamlss(y ~ trt + x, family = gamlss.dist::GA,
                        data = dm, trace = FALSE)
  ph <- gamlss_posthoc(fit, "trt", estimand = "mean", contrast = "pairwise",
                       comparison = "log_ratio", engine = "distribution",
                       uncertainty = "bootstrap", B = 39L, data = dm)
  c(Q = ph$contrasts$estimate[1], U = ph$contrasts$SE[1]^2)
}
est <- t(vapply(seq_len(M), one_fit, numeric(2)))
est <- as.data.frame(est)
est$imputation <- seq_len(M)
est
```

```{r pool, eval=has_mice && has_dist}
pool_rubin <- function(Q, U, nu_com) {
  M  <- length(Q)
  Qb <- mean(Q); Ub <- mean(U)
  B  <- stats::var(Q)
  T_ <- Ub + (1 + 1 / M) * B
  gamma <- (1 + 1 / M) * B / T_
  nu_old <- (M - 1) * (1 + Ub / ((1 + 1 / M) * B))^2
  nu_obs <- ((nu_com + 1) / (nu_com + 3)) * nu_com * (1 - gamma)
  nu_adj <- 1 / (1 / nu_old + 1 / nu_obs)
  tcrit <- stats::qt(0.975, df = nu_adj)
  data.frame(estimate = Qb, se = sqrt(T_),
             lower = Qb - tcrit * sqrt(T_),
             upper = Qb + tcrit * sqrt(T_),
             fmi = gamma, df = nu_adj)
}
pooled <- pool_rubin(est$Q, est$U, nu_com = nrow(d) - 3)
pooled
# Back to the ratio scale, which is where the result is reported
data.frame(ratio = exp(pooled$estimate),
           lower = exp(pooled$lower),
           upper = exp(pooled$upper))
```

```{r pool-plot, eval=has_gg && has_mice && has_dist}
pd <- data.frame(scale = "log-ratio", estimate = pooled$estimate,
                 lower = pooled$lower, upper = pooled$upper)
ggplot2::ggplot(pd, ggplot2::aes(x = estimate, y = scale)) +
  ggplot2::geom_vline(xintercept = 0, linetype = 2) +
  ggplot2::geom_errorbar(ggplot2::aes(xmin = lower, xmax = upper),
                         width = 0.15, orientation = "y") +
  ggplot2::geom_point(size = 2.5) +
  ggplot2::labs(x = "Pooled log-ratio, Bio versus Control", y = NULL,
                title = "Rubin-pooled post-hoc contrast") +
  theme_gamlss()
```

# 4. The scale on which to pool

Rubin's rules assume that \(\hat{Q}_m\) is approximately normal with a stable
within-imputation variance. That assumption is comfortable for a difference of
means and uncomfortable for the quantities distributional models produce most
often.

| Estimand | Pool on | Reason |
|---|---|---|
| difference of means | the difference itself | already symmetric |
| ratio, percent change | the log-ratio, then back-transform | ratios are bounded below by zero and right-skewed |
| variance, dispersion | the log | positive and right-skewed |
| probability, zero mass | the logit | bounded in the unit interval |
| quantile | the quantile itself, or the log for a positive response | usually mildly skewed |
| optimum located by root-finding | do not pool analytically | the sampling distribution can be multimodal |

For the last row, and for any estimand whose sampling distribution is clearly
non-normal, pool the *draws* rather than the summaries: collect the bootstrap or
simulation draws from every imputation into one pooled set and read the interval
off its quantiles. That approach requires `keep_draws = TRUE` and makes no
normality assumption at all.

Whatever is chosen, state it. "Pooled on the log scale and back-transformed" is
one clause in a methods section and removes an entire class of ambiguity.

# 5. What the missingness mechanism implies

| Mechanism | Definition | Consequence for the post-hoc analysis |
|---|---|---|
| MCAR | missingness independent of observed and unobserved data | complete-case analysis is unbiased but wasteful |
| MAR | missingness depends only on observed data | multiple imputation is valid, provided the imputation model includes the variables that drive missingness |
| MNAR | missingness depends on the unobserved value itself | no imputation model based on observed data is sufficient; sensitivity analysis is mandatory |

MAR is an assumption about the *imputation model*, not a property of the data.
It becomes more plausible as more of the variables predicting missingness are
included, which is why the imputation model should normally be richer than the
analysis model, including the outcome and the treatment indicator.

## 5.1 Sensitivity analysis under MNAR

The standard device is a delta adjustment. Shift the imputed values by a fixed
amount \(\delta\) on the analysis scale, repeat the whole pipeline, and report
how far \(\delta\) must depart from zero before the conclusion changes.

```{r mnar, eval=has_mice && has_dist}
delta_analysis <- function(delta) {
  vals <- vapply(seq_len(M), function(m) {
    dm <- mice::complete(imp, m)
    dm$x[is.na(d$x)] <- dm$x[is.na(d$x)] + delta
    fit <- gamlss::gamlss(y ~ trt + x, family = gamlss.dist::GA,
                          data = dm, trace = FALSE)
    gamlss_posthoc(fit, "trt", estimand = "mean", contrast = "pairwise",
                   comparison = "log_ratio", engine = "distribution",
                   uncertainty = "none", data = dm)$contrasts$estimate[1]
  }, numeric(1))
  mean(vals)
}
deltas <- c(-10, -5, 0, 5, 10)
data.frame(delta = deltas,
           pooled_log_ratio = round(vapply(deltas, delta_analysis, numeric(1)), 4))
```

If the sign and the practical size of the effect survive the plausible range of
\(\delta\), the conclusion is robust to a departure from MAR of that magnitude.
That is a much stronger statement than asserting MAR and moving on.

# 6. Missingness in the scale predictor

This case is specific to distributional regression and is almost absent from the
imputation literature. In a GAMLSS the same covariate can appear in
`mu.formula` and in `sigma.formula`. Two consequences follow.

First, the imputation model must be compatible with *both* roles. An imputation
model that preserves the conditional mean of `x` given the other variables but
destroys its association with the residual spread will bias the fitted dispersion
even if the mean structure is recovered correctly.

Second, the fraction of missing information is generally larger for a dispersion
estimand than for a mean estimand computed from the same imputations, because
second-order features are estimated less precisely. Reporting \(\gamma\)
separately for each estimand is therefore informative rather than redundant.

```{r fmi-by-estimand, eval=has_mice && has_dist}
fmi_for <- function(estimand) {
  z <- t(vapply(seq_len(M), function(m) {
    dm <- mice::complete(imp, m)
    fit <- gamlss::gamlss(y ~ trt + x, sigma.formula = ~ x,
                          family = gamlss.dist::GA, data = dm, trace = FALSE)
    ph <- gamlss_posthoc(fit, "trt", estimand = estimand, contrast = "pairwise",
                         comparison = "log_ratio", engine = "distribution",
                         uncertainty = "bootstrap", B = 29L, data = dm)
    c(ph$contrasts$estimate[1], ph$contrasts$SE[1]^2)
  }, numeric(2)))
  B <- stats::var(z[, 1]); Ub <- mean(z[, 2])
  T_ <- Ub + (1 + 1 / M) * B
  data.frame(estimand = estimand, fmi = round((1 + 1 / M) * B / T_, 4))
}
do.call(rbind, lapply(c("mean", "variance"), fmi_for))
```

# 7. What not to do

- Do not concatenate the imputed data sets and analyse them as one file. The
  sample size is inflated \(M\)-fold and every standard error is wrong.
- Do not impute the outcome and then discard the imputed outcome rows. That
  discards information and reintroduces selection.
- Do not pool a ratio directly and report a symmetric interval around it. The
  lower limit can fall below zero, which is impossible for a ratio.
- Do not report a pooled estimate without \(\gamma\). Without it the reader
  cannot tell whether the interval is driven by the data or by the missingness.

# References
