---
title: "A Data-First Graphics Layer for GAMLSS"
author: "Walter Esfrain Pereira"
output: rmarkdown::html_vignette
bibliography: references.bib
link-citations: true
vignette: >
  %\VignetteIndexEntry{A Data-First Graphics Layer for GAMLSS}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 6.2, fig.height = 3.8)
library(gamlssPosthoc)
has_core <- requireNamespace("gamlss", quietly = TRUE) &&
  requireNamespace("gamlss.dist", quietly = TRUE)
has_dist <- has_core && requireNamespace("distributions3", quietly = TRUE)
has_gg   <- has_core && requireNamespace("ggplot2", quietly = TRUE)
has_zadj <- has_dist && requireNamespace("gamlss.inf", quietly = TRUE)
has_broom <- requireNamespace("broom", quietly = TRUE)
has_flex <- requireNamespace("flextable", quietly = TRUE)
if (has_core) suppressPackageStartupMessages(library(gamlss.dist))
```

# 1. Introduction

A GAMLSS models every parameter of a conditional distribution [@rigby2005],

\[
Y_i \sim \mathcal{D}(\theta_{1i},\ldots,\theta_{pi}),\qquad
g_k(\theta_{ki}) = \eta_{ki}.
\]

Location, dispersion and shape can each depend on treatments, covariates and
smoothers. Version 0.3.0 adds a **data-first** graphics layer:

\[
\boxed{\text{model} \rightarrow \text{estimand or distribution}
\rightarrow \text{plot data} \rightarrow \text{ggplot}}
\]

The purpose is to separate the statistical computation from its visual
representation. `gamlss_plot_data()` produces the auditable table, and the
`plot_gamlss_*()` functions build `ggplot2` objects the user can modify. This
follows the public extension API of `ggplot2`, which recommends that packages
return `ggplot` objects and use `autoplot()` as the S3 mechanism for automatic
visualization [@wickham_ggplot2].

The practical consequence is that no figure in this package is a black box. Every
plot has a table behind it, that table is exported, and it can be checked,
re-plotted or shipped as supplementary material.

# 2. Why several parameters must be visualized

In a family with parameters \(\mu, \sigma, \nu, \tau\), treatments can act
differently on each:

\[
\mu = \mu(x),\quad \sigma = \sigma(x),\quad \nu = \nu(x),\quad \tau = \tau(x).
\]

A figure of \(\mu\) alone can hide important changes in dispersion or shape. The
package discovers the parameters present in the object dynamically, so the same
call works for a two-parameter and a four-parameter family.

# 3. Reproducible example

```{r fit, eval=has_core}
library(gamlssPosthoc)
set.seed(1001)
d <- data.frame(
  dose = rep(seq(0, 150, length.out = 12), each = 10),
  trt  = factor(rep(c("Control", "Bio"), length.out = 120))
)
mu <- exp(log(2.2) + .012 * d$dose - .00005 * d$dose^2 + .18 * (d$trt == "Bio"))
sigma <- exp(log(.22) + .002 * d$dose)
d$y <- gamlss.dist::rGA(nrow(d), mu = mu, sigma = sigma)
fit <- gamlss::gamlss(y ~ trt + dose + I(dose^2), sigma.formula = ~ dose,
                      family = gamlss.dist::GA, data = d, trace = FALSE)
```

## 3.1 Parameters

```{r params, eval=has_gg}
plot_gamlss_parameters(fit, "dose", parameters = c("mu", "sigma"), data = d)
```

Two selection styles are available, and both are executed below so the reader can
see the difference rather than read about it.

```{r params-all, eval=has_gg}
plot_gamlss_parameters(fit, x = "dose", parameters = "all", data = d)
```

## 3.2 Marginal estimand

The estimand must be declared. For the mean,

\[
T(F) = E(Y).
\]

```{r estimand, eval=has_gg && has_dist}
plot_gamlss_estimand(fit, "dose", estimand = "mean", data = d)
```

# 4. Predictive distributions

Visualizing the whole distribution exposes differences a mean cannot show. For a
standardized population,

\[
f_g(y) = \sum_i w_i\, f(y \mid X_i, g).
\]

```{r dist, eval=has_gg && has_dist}
plot_gamlss_distribution(fit, "trt", type = "density", data = d)
```

The cumulative and survival views use the same plot data.

```{r dist-cdf, eval=has_gg && has_dist}
plot_gamlss_distribution(fit, "trt", type = "cdf", data = d, n_response = 60L)
```

```{r dist-surv, eval=has_gg && has_dist}
plot_gamlss_distribution(fit, "trt", type = "survival", data = d, n_response = 60L)
```

# 5. Quantile fan, and why it is a mixture quantile

Marginal quantiles come from the standardized mixture distribution, not from
averaging the conditional quantiles. For \(p \in (0,1)\), the package solves

\[
\sum_i w_i F_i(q_p) = p
\tag{1}
\]

for \(q_p\). The naive alternative, \(\tilde q_p = \sum_i w_i F_i^{-1}(p)\), is a
different quantity. Because the quantile function is non-linear, Jensen's
inequality applies again, and the two coincide only when the conditional
distributions are identical across \(i\). The block below computes both.

```{r mixture-quantile, eval=has_dist}
pd <- gamlss_plot_data(fit, "quantiles", x = "dose", probs = 0.9,
                       data = d, n = 5L)
# Naive average of conditional quantiles at the same dose grid
grid <- sort(unique(pd$dose))
naive <- vapply(grid, function(g) {
  nd <- transform(d, dose = g)
  mu_i <- stats::predict(fit, what = "mu", newdata = nd, type = "response", data = d)
  sg_i <- stats::predict(fit, what = "sigma", newdata = nd, type = "response", data = d)
  if (length(sg_i) == 1L) sg_i <- rep(sg_i, length(mu_i))
  mean(gamlss.dist::qGA(0.9, mu = mu_i, sigma = sg_i))
}, numeric(1))
data.frame(dose = round(grid, 2),
           mixture_quantile = round(pd$quantile, 4),
           average_of_quantiles = round(naive, 4),
           relative_gap = round(pd$quantile / naive - 1, 4))
```

The gap is small here because the two treatment groups are similar at each dose,
and it widens as the conditional distributions separate. Equation (1) is the one
that answers "what value does 90 percent of this population fall below".

```{r fan, eval=has_gg && has_dist}
plot_gamlss_quantiles(fit, "dose", probs = c(.05, .10, .25, .50, .75, .90, .95),
                      style = "fan", data = d)
```

# 6. Zero-adjusted responses

For a semicontinuous response,

\[
P(Y = 0) = h, \qquad Y \mid Y > 0 \sim F_+ ,
\]

with marginal mean \(E(Y) = (1-h)E(Y_+)\) and variance

\[
\operatorname{Var}(Y) = (1-h)\operatorname{Var}(Y_+) + h(1-h)\{E(Y_+)\}^2 .
\]

`plot_gamlss_zero_adjusted()` shows the zero probability, the positive mean and
the marginal mean separately, which prevents `mu` from being read as if it
described the full response.

```{r zadj, eval=has_gg && has_zadj}
set.seed(1003)
dz <- data.frame(trt = factor(rep(c("A", "B"), each = 40)))
h <- ifelse(dz$trt == "A", .40, .18)
pos <- gamlss.dist::rGA(nrow(dz), mu = ifelse(dz$trt == "A", 2, 3), sigma = .30)
dz$y <- ifelse(stats::rbinom(nrow(dz), 1, h) == 1, 0, pos)
zfit <- gamlss.inf::gamlssZadj(y = y, mu.formula = ~ trt, xi0.formula = ~ trt,
                               family = gamlss.dist::GA, data = dz, trace = FALSE)
plot_gamlss_zero_adjusted(zfit, "trt", data = dz)
```

# 7. Scientific contrasts

The package separates *who* is compared from *how* the effect is expressed. For
groups A and B,

\[
\Delta = A - B, \qquad R = A/B, \qquad PC = 100(R - 1).
\]

```{r contrast, eval=has_gg && has_dist}
ph <- gamlss_posthoc(fit, specs = "trt", estimand = "mean", contrast = "pairwise",
                     comparison = "percent_change", population = "observed",
                     uncertainty = "none", data = d)
plot_gamlss_contrasts(ph)
```

With retained bootstrap draws and `ggdist` installed, `style = "distribution"`
renders the uncertainty distributionally rather than as an interval; `ggdist` is
an extension of the grammar of graphics built specifically for distributions and
uncertainty [@kay2023].

# 8. Trends, derivatives and optima

For \(m(x) = E(Y \mid x)\), the first derivative

\[
m'(x) = \frac{d\,m(x)}{dx}
\]

indicates local growth or decline, and the second describes curvature.

```{r trend, eval=has_gg && has_dist}
plot_gamlss_trend(fit, "dose", data = d)
```

```{r deriv, eval=has_gg && has_dist}
plot_gamlss_derivative(fit, "dose", order = 1, data = d)
```

```{r optimum, eval=has_gg && has_dist}
plot_gamlss_optimum(fit, "dose", optimum = "maximum", data = d)
```

A mathematical maximum should not be relabelled automatically as an economic or
recommended dose. That interpretation depends on costs, risks and other outcomes
the model does not contain.

# 9. Surfaces

When two quantitative predictors act on the estimand,

\[
T(F) = T(F \mid x_1, x_2),
\]

a surface can reveal response regions that one-dimensional slices hide.

```{r surface, eval=has_gg && has_dist}
set.seed(1004)
ds <- data.frame(dose = runif(90, 0, 150), salinity = runif(90, 0, 8))
ds$y <- gamlss.dist::rGA(nrow(ds),
  mu = exp(0.8 + 0.006 * ds$dose - 0.06 * ds$salinity), sigma = 0.30)
fit2 <- gamlss::gamlss(y ~ dose + salinity, sigma.formula = ~ salinity,
                       family = gamlss.dist::GA, data = ds, trace = FALSE)
plot_gamlss_surface(fit2, "dose", "salinity", estimand = "mean", data = ds, n = 14L)
```

```{r surface-var, eval=has_gg && has_dist}
plot_gamlss_surface(fit2, "dose", "salinity", estimand = "parameter",
                    what = "sigma", type = "contour", data = ds, n = 14L)
```

# 10. Observations against the predicted distribution

`plot_gamlss_fit()` overlays the observed data, the predicted median and a
quantile band. For continuous responses this is normally more informative than a
bar chart of means, which discards the entire distribution.

```{r fitplot, eval=has_gg && has_dist}
plot_gamlss_fit(fit, "dose", response = "y", probs = c(.05, .5, .95), data = d)
```

# 11. Diagnostics

GAMLSS relies on randomized quantile residuals [@dunn1996]. Under a correctly
specified model these are exactly standard normal, whatever the family, which is
what makes them comparable across candidate distributions.

```{r diag-qq, eval=has_gg}
plot_gamlss_diagnostics(fit, "qq", data = d)
```

```{r diag-worm, eval=has_gg}
plot_gamlss_diagnostics(fit, "worm", data = d)
```

```{r diag-pit, eval=has_gg}
plot_gamlss_diagnostics(fit, "pit", data = d)
```

The worm-style plot implemented here is a detrended quantile-quantile
representation in `ggplot2`; the native GAMLSS diagnostics remain complementary.
The former `gamlss.ggplots` package documented a wide collection of GAMLSS plots
in `ggplot2` but was archived on CRAN in November 2025. `gamlssPosthoc`
therefore does not depend on it, and implements its diagnostics with public
`ggplot2` APIs only.

# 12. Model comparison

```{r compare, eval=has_gg}
fit_ga <- gamlss::gamlss(y ~ trt + dose, family = gamlss.dist::GA,
                         data = d, trace = FALSE)
fit_no <- gamlss::gamlss(y ~ trt + dose, family = gamlss.dist::NO,
                         data = d, trace = FALSE)
plot_gamlss_compare(GA = fit_ga, NO = fit_no, type = "criteria")
```

```{r compare-estimand, eval=has_gg && has_dist}
plot_gamlss_compare(GA = fit_ga, NO = fit_no, type = "estimand", x = "dose",
                    data = d)
```

AIC and GAIC must be read together with diagnostics and scientific plausibility.
A figure does not replace family selection or external validation; see
`vignette("advanced-workflows")` for proper scoring rules.

# 13. The separate data layer

Any figure can be rebuilt by the user from its table.

```{r datafirst, eval=has_dist}
pd <- gamlss_plot_data(fit, type = "parameters", x = "dose",
                       parameters = c("mu", "sigma"), data = d)
head(pd)
```

```{r datafirst-custom, eval=has_gg && has_dist}
ggplot2::ggplot(pd, ggplot2::aes(dose, estimate)) +
  ggplot2::geom_line(linewidth = 0.9) +
  ggplot2::facet_wrap(~ parameter, scales = "free_y") +
  ggplot2::labs(x = "Dose", y = "Parameter value",
                title = "Rebuilt by the user from the exported plot data") +
  theme_gamlss()
```

This layer matters for reproducibility, for regression tests, and for journals
that require the numbers behind every figure.

# 14. `autoplot()`

`ggplot2::autoplot()` is the official S3 mechanism for producing a complete plot
appropriate to a class. Methods are provided for `gamlss_posthoc`,
`gamlss_trend`, `gamlss_distribution_summary` and `gamlss_posthoc_plan`.

```{r autoplot, eval=has_gg && has_dist}
ggplot2::autoplot(ph)
```

# 15. Interoperability

Predictions, comparisons, slopes and marginal quantities should be described
explicitly as transformations of the model rather than as properties of it
[@arelbundock2024]. That philosophy is adopted here, and the result objects
implement the standard extraction generics.

```{r broom, eval=has_broom && has_dist}
head(generics::tidy(ph))
generics::glance(ph)
head(generics::augment(ph))
```

# 16. Export and reports

```{r export, eval=has_dist}
cat(substr(export_to_latex(ph, component = "contrasts"), 1, 160))
```

```{r flex, eval=has_flex && has_dist}
as_flextable(ph, component = "contrasts")
```

```{r report, eval=has_dist}
cat(substr(generate_report(ph, output = "md", include_plots = FALSE), 1, 160))
```

# 17. Themes

The themes return ordinary `ggplot2::theme` objects and can be overridden.

```{r themes, eval=has_gg && has_dist}
p <- plot_gamlss_estimand(fit, "dose", estimand = "mean", data = d)
p + theme_gamlss()
```

```{r themes2, eval=has_gg && has_dist}
p + theme_gamlss_journal()
```

```{r themes3, eval=has_gg && has_dist}
p + theme_gamlss_presentation()
```

# 18. Accessibility and journal requirements

Three practical constraints recur in submission, and all three are handled at the
`ggplot2` level rather than inside the package, precisely because the returned
object is an ordinary plot.

**Colour.** Roughly eight percent of men have some form of colour vision
deficiency, so a figure must not encode its message in hue alone. Use a
colour-blind-safe palette and add a redundant channel, such as line type.

```{r accessible, eval=has_gg && has_dist}
plot_gamlss_estimand(fit, "dose", estimand = "mean", by = "trt", data = d) +
  ggplot2::scale_colour_viridis_d(option = "cividis", end = 0.85) +
  theme_gamlss_journal()
```

**Resolution and format.** Most journals require vector formats for line art and
at least 300 dpi for raster. Both are one call away:

```{r export-figure, eval=FALSE}
ggplot2::ggsave("figure1.eps", p, width = 88, height = 60, units = "mm")
ggplot2::ggsave("figure1.tiff", p, width = 88, height = 60, units = "mm",
                dpi = 600, compression = "lzw")
```

**Size and type.** `theme_gamlss_journal()` defaults to a small base size suited
to a single-column figure. Set `base_family` to the journal's required font
rather than rescaling the exported image, which distorts the type.

# 19. Good practice

1. Declare the estimand before choosing the figure.
2. Do not treat `mu` automatically as \(E(Y)\).
3. For zero-adjusted responses, present the zero mass and the positive part when
   both are scientifically relevant.
4. Prefer estimates with intervals over isolated group letters.
5. Use the bootstrap for quantities derived from several parameters when the
   joint covariance is not available.
6. Do not extrapolate derivatives or optima beyond the support of the data.
7. Audit the engine with `gamlss_posthoc_plan()` before inference.
8. Ship the plot data alongside the figure.

# 20. Verified software documentation

- `ggplot2::autoplot()`: https://ggplot2.tidyverse.org/reference/autoplot.html
- Using `ggplot2` in packages: https://ggplot2.tidyverse.org/articles/ggplot2-in-packages.html
- `ggdist`: https://mjskay.github.io/ggdist/
- `marginaleffects`: https://marginaleffects.com/
- `gamlss.ggplots` archive record: https://CRAN.R-project.org/package=gamlss.ggplots
- `vdiffr::expect_doppelganger()`: https://vdiffr.r-lib.org/reference/expect_doppelganger.html

# References
