---
title: "Estimands, Standardization, and Contrasts after GAMLSS"
author: "Walter Esfrain Pereira"
output: rmarkdown::html_vignette
bibliography: references.bib
link-citations: true
vignette: >
  %\VignetteIndexEntry{Estimands, Standardization, and Contrasts after GAMLSS}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(gamlssPosthoc)
has_core <- requireNamespace("gamlss", quietly = TRUE) &&
  requireNamespace("gamlss.dist", quietly = TRUE)
has_dist <- has_core && requireNamespace("distributions3", quietly = TRUE)
has_zadj <- has_dist && requireNamespace("gamlss.inf", quietly = TRUE)
has_emm  <- has_core && requireNamespace("emmeans", quietly = TRUE)
# gamlss.inf resolves distribution functions such as pGG() by name from the
# search path, so gamlss.dist must be attached, not merely loaded.
if (has_zadj) suppressPackageStartupMessages(library(gamlss.dist))
```

## Five separate decisions

A post-hoc analysis after GAMLSS should state five decisions explicitly:

1. **Estimand:** a distributional parameter, the mathematical response mean,
   variance, a quantile, the exact zero mass, or another function.
2. **Target population:** observed, balanced, or a reference profile.
3. **Comparison geometry:** pairwise, reference, sequential, or polynomial.
4. **Scientific scale:** difference, ratio, log-ratio, or percent change.
5. **Uncertainty layer:** delta, coefficient simulation, refit bootstrap, or point estimate only.

`gamlssPosthoc` stores these choices in the returned object instead of using the
ambiguous phrase "adjusted means" for all of them.

## What an estimand is, formally

A GAMLSS specifies a full conditional distribution [@rigby2005],

\[
Y_i \sim \mathcal{D}(\theta_{1i},\ldots,\theta_{pi}),\qquad
g_k(\theta_{ki}) = \eta_{ki} = \mathbf{x}_{ki}^{\top}\boldsymbol{\beta}_k
+ \sum_j s_{kj}(x_{kji}).
\]

Every quantity this package reports is a functional \(T\) of a *standardized*
predictive distribution. Fix a level \(g\) of the focal factor, keep the
remaining covariates at their observed values, and average the conditional
distributions,

\[
F_g(y) \;=\; \sum_{i=1}^{n} w_i\, F\!\left(y \mid \mathbf{X}_i,\, g\right),
\qquad w_i \ge 0,\quad \sum_i w_i = 1 .
\tag{1}
\]

The estimand is then \(T(F_g)\). The package options map onto \(T\) as follows.

| `estimand` | Functional \(T(F)\) |
|---|---|
| `parameter` | \(\theta_k\), a coordinate of the parameter vector, not a functional of \(F\) as a whole |
| `mean` | \(\int y \, dF(y)\) |
| `variance` | \(\int y^2 dF(y) - \left(\int y\, dF(y)\right)^2\) |
| `quantile` | \(F^{-1}(p)\), solved from \(\sum_i w_i F_i(q_p) = p\) |
| `prob_zero` | \(F(0) - F(0^-)\), the exact point mass |
| `custom` | any user-supplied \(T\) |

This is why `estimand` and `what` are separate arguments. `what` selects a
coordinate \(k\); `estimand` selects the functional. Reporting "the mean" and
"mu" as if they were the same quantity is the single most common error in
applied GAMLSS post-hoc work, and under a non-identity link they are not equal.

## Why the mean is not the back-transformed linear predictor

Under a log link, \(\mu_i = \exp(\eta_i)\). Because the exponential is strictly
convex, Jensen's inequality gives

\[
\mathrm{E}\!\left[g^{-1}(\eta)\right] \;>\; g^{-1}\!\left(\mathrm{E}[\eta]\right)
\]

whenever \(\eta\) varies across the population being averaged. The gap is not a
rounding artifact; it grows with the dispersion of \(\eta\). The following block
computes both quantities on the same fit.

```{r jensen, eval=has_core}
set.seed(2026)
dj <- data.frame(
  trt  = factor(rep(c("Control", "A"), each = 40)),
  dose = rep(seq(0, 120, length.out = 40), 2)
)
mj <- exp(0.7 + 0.010 * dj$dose + 0.30 * (dj$trt == "A"))
dj$y <- gamlss.dist::rGA(nrow(dj), mu = mj, sigma = 0.35)
fj <- gamlss::gamlss(y ~ trt + dose, family = gamlss.dist::GA,
                     data = dj, trace = FALSE)

# Standardized mean of the back-transformed predictions, which is equation (1)
nd_a <- transform(dj, trt = factor("A", levels = levels(dj$trt)))
nd_c <- transform(dj, trt = factor("Control", levels = levels(dj$trt)))
m_a <- mean(stats::predict(fj, what = "mu", newdata = nd_a,
                           type = "response", data = dj))
m_c <- mean(stats::predict(fj, what = "mu", newdata = nd_c,
                           type = "response", data = dj))

# Back-transformation of the averaged linear predictor, which is not the mean
e_a <- exp(mean(stats::predict(fj, what = "mu", newdata = nd_a,
                               type = "link", data = dj)))
e_c <- exp(mean(stats::predict(fj, what = "mu", newdata = nd_c,
                               type = "link", data = dj)))

data.frame(level = c("Control", "A"),
           standardized_mean = c(m_c, m_a),
           back_transformed  = c(e_c, e_a),
           relative_gap = c(m_c / e_c - 1, m_a / e_a - 1))
```

The two columns differ, and the difference propagates to any contrast built on
top of them. `gamlssPosthoc` always reports the first column.

## Mapping to the ICH E9(R1) estimand attributes

The regulatory estimand framework [@ich2019] has become the common vocabulary
for describing what a number means before describing how it was computed. The
correspondence with the package arguments is direct.

| ICH E9(R1) attribute | `gamlssPosthoc` argument |
|---|---|
| Population | `population`, `weights`, `grid` |
| Variable, or endpoint | `estimand`, `what`, `prob`, `custom_fun` |
| Treatment condition | `specs`, `at` |
| Population-level summary | `contrast`, `comparison` |
| Intercurrent event handling | analyst's responsibility, recorded in `at` and in the fitted model |

Writing these five attributes into a methods section is enough to make a GAMLSS
post-hoc analysis reproducible by a reader who does not have the code.

## Gamma example and the diagnostic plan

```{r gamma-example, eval=has_core}
set.seed(2026)
d <- data.frame(trt = factor(rep(c("Control", "A", "B"), each = 35)))
mu <- c(Control = 2.0, A = 2.6, B = 3.1)[d$trt]
d$y <- gamlss.dist::rGA(nrow(d), mu = mu, sigma = 0.35)
fit <- gamlss::gamlss(y ~ trt, family = gamlss.dist::GA,
                      data = d, trace = FALSE)

gamlss_posthoc_plan(fit, estimand = "parameter", what = "mu")
```

The plan reports model class, family, discovered parameters, smooth terms,
optional package availability, a per-parameter capability matrix, engine
eligibility, and a recommended inference route.

For a simple parameter-wise request, `emmeans` [@lenth_emmeans] is selected
automatically only when the target is explicitly `population = "reference"`.
This prevents an observed-population estimand from being silently replaced by an
EMM reference grid. Observed or balanced standardization is routed through
prediction-based engines [@arelbundock2024].

## The estimand is explicit

The response mean can be standardized over the empirical covariate distribution:

```{r mean-example, eval=has_dist}
ans <- gamlss_posthoc(
  fit, specs = "trt", estimand = "mean",
  population = "observed",
  contrast = "reference",
  comparison = "percent_change",
  uncertainty = "none",
  data = d
)
ans$estimand_info
ans$contrasts
```

## The comparison scale is a separate decision

The comparison scale is independent of which levels are compared. Let
\(\bar{T}_g = T(F_g)\) denote the standardized estimand at level \(g\). For two
levels \(a\) and \(b\), the four supported scales are

\[
\begin{aligned}
\text{difference} &: \bar{T}_a - \bar{T}_b, \\
\text{ratio} &: \bar{T}_a / \bar{T}_b, \\
\text{log-ratio} &: \log\bar{T}_a - \log\bar{T}_b, \\
\text{percent change} &: 100\left(\bar{T}_a / \bar{T}_b - 1\right).
\end{aligned}
\]

Two consequences deserve emphasis. First, the ratio is computed from the two
standardized estimands, so it is a ratio of averages and not the average of the
row-wise ratios; those two quantities are different whenever the covariate
distribution is not degenerate. Second, the log-ratio is the only one of the
four that is approximately symmetric and approximately normal in small samples,
which makes it the safer scale for pooling and for meta-analysis.

```{r scales, eval=has_dist}
scales <- c("difference", "ratio", "log_ratio", "percent_change")
do.call(rbind, lapply(scales, function(s) {
  z <- gamlss_posthoc(fit, "trt", estimand = "mean", contrast = "reference",
                      comparison = s, uncertainty = "none", data = d)$contrasts
  data.frame(scale = s, comparison = z$comparison, estimate = round(z$estimate, 4))
}))
```

## Generic zero-adjusted distributions

For a zero-adjusted distribution with zero probability \(h\) and positive-part
random variable \(Y_+\), the full response has

\[
E(Y) = (1-h)E(Y_+)
\]

and

\[
\operatorname{Var}(Y) = (1-h)\operatorname{Var}(Y_+)
+ h(1-h)\{E(Y_+)\}^2.
\]

The variance formula is the law of total variance applied to the Bernoulli
indicator of a positive observation, and the second term is exactly the
between-component contribution that an analysis restricted to positive values
would discard.

The package reconstructs the positive distribution through
`gamlss.dist::GAMLSS()` whenever possible [@rigby2019]. Therefore, a generalized
Gamma or another compatible positive family no longer needs a hand-written
formula for its mean, variance, or quantiles.

```{r zadj-example, eval=has_zadj}
set.seed(2027)
z <- data.frame(trt = factor(rep(c("Control", "Bio"), each = 55)))
zmu <- ifelse(z$trt == "Control", 2.0, 3.0)
zh <- ifelse(z$trt == "Control", .40, .18)
zpos <- gamlss.dist::rGG(nrow(z), mu = zmu, sigma = .35, nu = .7)
z$y <- ifelse(stats::rbinom(nrow(z), 1, zh) == 1, 0, zpos)
zfit <- gamlss.inf::gamlssZadj(
  y = y, mu.formula = ~ trt, xi0.formula = ~ trt,
  family = gamlss.dist::GG, data = z, trace = FALSE
)
nd <- data.frame(trt = factor(levels(z$trt), levels = levels(z$trt)))
gamlss_distribution_summary(zfit, nd, data = z)
```

For a locally generated family that `gamlss.dist::GAMLSS()` cannot reconstruct,
provide `positive_dist_fun` rather than separate ad-hoc formulas for each
moment.

## Layered uncertainty

Let \(\hat{\boldsymbol{\beta}}\) be the fitted coefficient vector with estimated
covariance \(\hat{V}\), and let the estimand be \(\psi = h(\boldsymbol{\beta})\).
The delta method uses the first-order expansion

\[
\operatorname{Var}(\hat\psi) \;\approx\;
\nabla h(\hat{\boldsymbol{\beta}})^{\top}\,\hat{V}\,
\nabla h(\hat{\boldsymbol{\beta}}) .
\tag{2}
\]

Equation (2) is trustworthy when \(h\) is smooth and close to linear over the
region where \(\hat{\boldsymbol{\beta}}\) has appreciable mass. For a
distributional parameter on the link scale that condition normally holds. For a
functional of the whole predictive distribution, such as a marginal variance, a
mixture quantile, or an optimum located by root-finding, it often does not, and
in current `gamlss` there is additionally no reliable joint covariance interface
across all parameter blocks. The package therefore refuses to fabricate a delta
interval in that situation and falls back to a refit bootstrap.

The layers are:

- `delta`: covariance-based inference through equation (2), where the selected
  engine supports it;
- `simulation`: coefficient simulation through `marginaleffects` when available,
  which relaxes the linearity requirement but keeps the normal approximation for
  \(\hat{\boldsymbol{\beta}}\);
- `bootstrap`: model refitting for generic derived targets, the only layer that
  makes no local-linearity assumption;
- `none`: plug-in point estimates.

The `auto` mode avoids silently launching hundreds of model refits. Use
`gamlss_posthoc_plan()` to see when an explicit refit bootstrap is recommended.

## Quantitative predictors

`gamlss_trend()` is intentionally different from a polynomial contrast among
factor levels. For a numeric dose it can report the curve, first or second
numerical derivatives, turning points, and optima refined by a local quadratic
interpolation when the grid extremum is internal and has compatible curvature.

Derivatives are computed by central differences on the standardized estimand,

\[
\widehat{T}'(x) \approx \frac{\bar{T}(x+\delta) - \bar{T}(x-\delta)}{2\delta},
\]

so they inherit the target population of the estimand itself rather than
describing a single covariate profile.

```{r quantitative, eval=has_dist}
set.seed(2028)
qd <- data.frame(dose = seq(0, 10, length.out = 100))
qd$y <- gamlss.dist::rGA(nrow(qd),
  mu = exp(1 + .20 * qd$dose - .020 * qd$dose^2), sigma = .20)
qfit <- gamlss::gamlss(y ~ dose + I(dose^2), family = gamlss.dist::GA,
                       data = qd, trace = FALSE)
g <- seq(.2, 9.8, length.out = 50)

gamlss_trend(qfit, "dose", at_x = g, method = "turning_points",
             estimand = "mean", uncertainty = "none", data = qd)$special_points
```

When a simultaneous band is requested, the package rescales the pointwise
standard errors by a critical value obtained from the maximum absolute
studentized deviation across the grid,

\[
c_{1-\alpha} = \text{quantile}_{1-\alpha}\left(
\max_{x \in \mathcal{G}} \left| \frac{T^{(b)}(x) - \bar{T}(x)}{\widehat{se}(x)} \right|
\right),
\]

which controls coverage over the whole curve rather than at each point
separately. A pointwise band is narrower and answers a different question.

First derivatives of an ordinary distributional parameter can also be delegated
to `marginaleffects::avg_slopes()` when that package and the model interface are
available. Higher derivatives and full-distribution targets remain
prediction-based.

## Where to go next

- `vignette("estimands-and-standardization")` develops the target population
  formally and shows how much a conclusion can move between the three choices.
- `vignette("case-study-agronomy")` runs one complete analysis, from design to a
  written Results paragraph.
- `vignette("cookbook")` gives three worked calls for every exported function.

## References
