---
title: "Estimands, Standardization, and Contrasts after GAMLSS"
author: "Walter Esfrain Pereira"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Estimands, Standardization, and Contrasts after GAMLSS}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(gamlssPosthoc)
has_core <- requireNamespace("gamlss", quietly = TRUE) &&
  requireNamespace("gamlss.dist", quietly = TRUE)
has_dist <- has_core && requireNamespace("distributions3", quietly = TRUE)
has_zadj <- has_dist && requireNamespace("gamlss.inf", quietly = TRUE)
# `gamlss.inf::gamlssZadj()` resolves the density and CDF of the positive
# family by name with `eval(parse(text = ...))`, so gamlss.dist must be
# attached rather than merely namespace-loaded.
if (has_core) suppressMessages(library(gamlss.dist))
```

## Five separate decisions

A post-hoc analysis after GAMLSS should state five decisions explicitly:

1. **Estimand:** a distributional parameter, the mathematical response mean,
   variance, a quantile, the exact zero mass, or another function.
2. **Target population:** observed, balanced, or a reference profile.
3. **Comparison geometry:** pairwise, reference, sequential, or polynomial.
4. **Scientific scale:** difference, ratio, log-ratio, or percent change.
5. **Uncertainty layer:** delta, coefficient simulation, refit bootstrap, or point estimate only.

`gamlssPosthoc` stores these choices in the returned object instead of using the
ambiguous phrase "adjusted means" for all of them.

## Gamma example and the diagnostic plan

```{r gamma-example, eval=has_core}
set.seed(2026)
d <- data.frame(trt = factor(rep(c("Control", "A", "B"), each = 35)))
mu <- c(Control = 2.0, A = 2.6, B = 3.1)[d$trt]
d$y <- gamlss.dist::rGA(nrow(d), mu = mu, sigma = 0.35)
fit <- gamlss::gamlss(y ~ trt, family = gamlss.dist::GA,
                      data = d, trace = FALSE)

gamlss_posthoc_plan(fit, estimand = "parameter", what = "mu")
```

The plan reports model class, family, discovered parameters, smooth terms,
optional package availability, a per-parameter capability matrix, engine
eligibility, and a recommended inference route.

For a simple parameter-wise request, `emmeans` is selected automatically only
when the target is explicitly `population = "reference"`. This prevents an
observed-population estimand from being silently replaced by an EMM reference
grid. Observed or balanced standardization is routed through prediction-based
engines.

## The estimand is explicit

The response mean can be standardized over the empirical covariate distribution:

```{r mean-example, eval=has_dist}
ans <- gamlss_posthoc(
  fit, specs = "trt", estimand = "mean",
  population = "observed",
  contrast = "reference",
  comparison = "percent_change",
  uncertainty = "none",
  data = d
)
ans$estimand_info
ans$contrasts
```

The comparison scale is independent of which levels are compared. Replacing
`comparison = "percent_change"` by `"difference"`, `"ratio"`, or
`"log_ratio"` changes the scientific effect without changing the reference
geometry. For standardized marginal means, percent change is defined as
`100 * (mean_hi / mean_lo - 1)`; it is not the average of row-specific percent
changes.

## Generic zero-adjusted distributions

For a zero-adjusted distribution with zero probability \(h\) and positive-part
random variable \(Y_+\), the full response has

\[
E(Y) = (1-h)E(Y_+)
\]

and

\[
\operatorname{Var}(Y) = (1-h)\operatorname{Var}(Y_+)
+ h(1-h)\{E(Y_+)\}^2.
\]

Version 0.2.0 reconstructs the positive distribution through
`gamlss.dist::GAMLSS()` whenever possible. Therefore, a generalized Gamma or
another compatible positive family no longer needs a hand-written formula for
its mean, variance, or quantiles.

```{r zadj-example, eval=has_zadj}
set.seed(2027)
z <- data.frame(trt = factor(rep(c("Control", "Bio"), each = 55)))
zmu <- ifelse(z$trt == "Control", 2.0, 3.0)
zh <- ifelse(z$trt == "Control", .40, .18)
zpos <- gamlss.dist::rGG(nrow(z), mu = zmu, sigma = .35, nu = .7)
z$y <- ifelse(stats::rbinom(nrow(z), 1, zh) == 1, 0, zpos)
zfit <- gamlss.inf::gamlssZadj(
  y = y, mu.formula = ~ trt, xi0.formula = ~ trt,
  family = gamlss.dist::GG, data = z, trace = FALSE
)
nd <- data.frame(trt = factor(levels(z$trt), levels = levels(z$trt)))
gamlss_distribution_summary(zfit, nd, data = z)
```

For a locally generated family that `gamlss.dist::GAMLSS()` cannot reconstruct,
provide `positive_dist_fun` rather than separate ad-hoc formulas for each
moment.

## Layered uncertainty

The package supports several layers:

- `delta`: native covariance-based inference where the selected engine supports it;
- `simulation`: coefficient simulation through `marginaleffects` when available;
- `bootstrap`: model refitting for generic derived targets;
- `none`: plug-in point estimates.

The `auto` mode avoids silently launching hundreds of model refits. Use
`gamlss_posthoc_plan()` to see when an explicit refit bootstrap is recommended.

## Quantitative predictors

`gamlss_trend()` is intentionally different from a polynomial contrast among
factor levels. For a numeric dose it can report the curve, first or second
numerical derivatives, turning points, and optima refined by a local quadratic interpolation when the grid extremum is internal and has compatible curvature.

```{r quantitative, eval=has_dist}
set.seed(2028)
qd <- data.frame(dose = seq(0, 10, length.out = 100))
qd$y <- gamlss.dist::rGA(nrow(qd),
  mu = exp(1 + .20 * qd$dose - .020 * qd$dose^2), sigma = .20)
qfit <- gamlss::gamlss(y ~ dose + I(dose^2), family = gamlss.dist::GA,
                       data = qd, trace = FALSE)
g <- seq(.2, 9.8, length.out = 50)

gamlss_trend(qfit, "dose", at_x = g, method = "turning_points",
             estimand = "mean", uncertainty = "none", data = qd)$special_points
```

First derivatives of an ordinary distributional parameter can also be delegated
to `marginaleffects::avg_slopes()` when that package and the model interface are
available. Higher derivatives and full-distribution targets remain
prediction-based.
